---
title: 'Project Management Foundations'
author: "Zehui Bai"
date: '`r format(Sys.time())`'
output:
  html_document:
    df_print: paged
    number_sections: yes
    toc: yes
    toc_float: yes
  word_document:
    toc: yes
  pdf_document:
    toc: yes
fontsize: 10pt
editor_options:
  chunk_output_type: console
colorlinks: yes
---

```{r setup, include=FALSE, echo = FALSE,message = FALSE, error = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)

# <!-- ---------------------------------------------------------------------- -->
# <!--                    1. load the required packages                       -->
# <!-- ---------------------------------------------------------------------- --> 

## if(!require(psych)){install.packages("psych")}

packages<-c("tidyverse", "knitr", "papeR")
ipak <- function(pkg){
  new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
  if (length(new.pkg)) 
    install.packages(new.pkg, dependencies = TRUE)
  sapply(pkg, require, character.only = TRUE)
}
ipak(packages)
# <!-- ---------------------------------------------------------------------- --> 


# <!-- ---------------------------------------------------------------------- -->
# <!--                        2. Basic system settings                        -->
# <!-- ---------------------------------------------------------------------- -->
setwd(dirname(rstudioapi::getSourceEditorContext()$path))
getwd()
Sys.setlocale("LC_ALL","English")
```



```{r mind map,echo = F,message = FALSE, error = FALSE, warning = FALSE}
## Convert to mind map text, markdown outline, R script, and HTML widget ####
library(mindr)
# text -> widget
# input <- c("# Chapter 1", "## Section 1.1", "### Section 1.1.1", "## Section 1.2", "# Chapter 2")
# mm(from = input, root = "mindr")


input <- rstudioapi::getSourceEditorContext()$path 
## file.show(input) # Open the input file with the default program, if any
input_txt <- readLines(input, encoding = "UTF-8")
## Convert to mind map text, markdown outline, R script, and HTML widget ####
markmap(input_txt, root = "Fundamentals of digital marketing")
```

# Get to Know Project Management


## What is a project?

A project is a temporary endeavor that has a unique goal, and usually a budget. That's a start, but what does it really mean? Let's begin with temporary endeavor. Unlike day to day operations, a project has a definite beginning and end. If a project is implementing a new system, the project is done when you successfully hand it over to operations. What if a project seems to go on forever? It could be that you haven't clearly defined what you're trying to accomplish. That brings us to the project goal. A project produces a unique result, which could be a product, service, or other outcome. In the case study we're using for this course, a hospital is implementing a new scheduling system. Lots of organizations implement scheduling systems, but each one has its own set of objectives, constraints, and issues. Maybe the hospital struggles with staff shortages. Or perhaps a big pain point is scheduling equipment along with the skilled staff and rooms needed to operate it. You could be upgrading the scheduling system to improve results and level of care. Finally, most projects have budgets. Most of the time, you think money when you hear the word budget. In projects, you'll probably face other constraints as well, like how many resources are available. A project isn't the same as operations, which represents work that's the same day after day, producing the same results. Admitting patients into the hospital represents operational work. Patients may show up with different medical issues, but the people at the front desk follow the same procedures, and use the same forms to admit them. Introducing a new scheduling system into the hospital, on the other hand, is a project. It will have a specific beginning, it ends when the system is up and running, scheduling hospital resources. It has a unique goal to solve the hospital's scheduling issues, and it has a budget, both in money and resources time. Over your lifetime, you've probably worked on lots of projects, both at work and at home. A project is a temporary endeavor with a unique goal, and usually a budget. Using those characteristics, think about what you've done over the past several months. Identify the projects you've worked on during that time by identifying the project's timeframe, what made the project unique, and its budget. In the exercise files, you'll find a list of the work I've done recently, tagged either as projects or operations.

## Project management
 
Project management can be summed up as answering a few crucial questions. The first question you have to answer is, what problem are you solving? Clearly defining what the project is supposed to accomplish is a big step toward making it a success. Legendary baseball manager Yogi Berra's off-kilter take on it is, "You've got to be very careful "if you don't know where you're going, "because you might not get there." The second question is, how are you going to solve this problem? Whether you're solving a problem, or pursuing an opportunity, you might have to choose from several possible strategies. Once you've picked your approach, it's time to flesh out your solution, gathering requirements, identifying deliverables, and defining project scope. The next question is, what's your plan? A big part of project management is planning your project. You have to identify the work to be done, in detail. How long it might take, the resources you need, and how much they cost. With that info in hand, you can build a schedule of when work should occur. While you're at it, you also need to spell out how you want things to happen in your project, like communication, managing changes, and so on. Some projects seem to go on forever, but eventually someone will pull the plug if it doesn't finish. That's why you also have to answer the question, how will you know when you're done? Clearly defined objectives, requirements, and deliverables help answer that question, but you can eliminate uncertainty by defining success criteria, quantifiable, measurable results that show that the project is complete, like a certificate of occupancy, for construction, or increased productivity measurements after a system is implemented. When you get to the end of the project, you're ready to answer the last question, how well did the project go? This important step is often skipped, because everyone is so ready to move on to the next deal. You really need to take time to review the project. What worked well, what didn't, why? How could we have done better? Project management boils down to answering several questions about your project, what problem are you solving, how are you going to solve it, what's your plan for getting the project done, how can you tell when you're done, and how well did the project go?


## What it takes to be a project manager 

Most people see project managers as really organized and good at getting things done. But project managers also use a wide variety of skills and knowledge to complete projects successfully. First, there are technical skills specific to project management, like what goes into a project plan, building and fine tuning a project schedule, reading a Gant chart, using the critical path and measuring performance. Then there's business expertise. It's up to you to make sure your project delivers value. You need to understand your organization's business, what it does, and what it considers important. That way, you can see how your project fits in and you can make good decisions to ensure that your project is successful. Problem solving is another important skill. Projects never, and I mean never, go according to plan. Your job is to figure out how to achieve project objectives and meet the schedule and budget while working around the problems that arise. One of the most important things in a project manager's toolbox is interpersonal skills. Projects may use people from different groups, departments, and even different companies. You're the leader of this group so it's your job to help everyone work as a team to get the project done, and done well. The most important characteristic of a great project manager is strong leadership. You want to inspire your people, help them become a true team, guide them to do the right things, hold them accountable, and motivate them to give their best. Does project management sound like something you want to do? This course will help you figure out what you need to know so you can build a plan for developing your skills.

## Üroject management life cycle

Project management can be categorized into five groups of processes that help guide a project successfully from beginning to end. Here are project management's five process groups according to the Project Management Institute. Initiating is all about getting the commitment to start a project. To get that commitment, you start by defining the project. What's the project supposed to accomplish? What's the scope? What's a rough estimate of the resources needed, and the cost? You also identify the project stakeholders, and make sure they agree on what the project is. From there, you ask for approval to proceed. Planning is where you figure out how you're going to perform the project. In essence, planning answers the questions, what are we going to do? How are we going to do it? And how will we know when we're done? When the plan is complete, it's time to get approval to launch the project. The next two areas involve putting your plan into action. Executing starts with launching a project. You bring your resources on board, get them settled in, and explain the rules you're using to run the project. After that, everyone jumps in to put the plan into action. Monitoring and controlling a project means checking, what's going on in the project, and how that compares to what you planned. If the project is sliding off track, you take action to get it back on course. Finally, there's the closing process. This part is short, but important. You get the client to officially accept that the project is complete. You document the project performance, gather lessons learned, close contracts, and help resources move on to their next assignments. Those are the five process groups that represent the project management life cycle. Put these processes into action consistently, and your projects will run more smoothly.


## Traditional vs. agile project management 

Let's examine two common approaches to project management, and how to choose the right one for your projects. When each project management process group occurs one after another, it's known as traditional project management, or the waterfall approach, because the processes flow from one to the next. Waterfall project management works well when the project goal and solution are clearly defined, and the scope and deliverables are clear-cut. Because you understand what needs to be done, you step through each process group once from start to finish. The more you know about the project, the better the waterfall approach works. Simple projects with very little uncertainty are great candidates, because you know what needs to be done and how to handle issues that arise. With familiar technology, your team can be productive, because they know the potential problems and how to work around them. If teams have worked on similar projects in the past, they can be more productive because they understand the work and know how to prevent or resolve common problems. With many projects today, you don't know what the solution looks like, so you have to figure it out as you go. This type of project requires a different approach. Agile project management goes through iterations, sometimes called sprints, to deliver partial, yet production-quality solutions at regular intervals. With this approach, the customer gets value from the project sooner. In addition, the customer's feedback on what's been delivered so far can help improve the overall solution. The customer has to be more involved than in traditional projects. Project teams tend to be smaller, more experienced, and able to work without much supervision. During initiating and planning, you define the overall goal for the project and build a plan to achieve that goal. With agile project management, you then develop detailed plans for each iteration. Executing is easier because you work on one iteration at a time. And small independent teams of highly-skilled people make it easier to get the work done. Another characteristic of agile projects is that you monitor and control them more closely, and teams tend to communicate faster and more frequently. Finally, each iteration has its own closing process for accepting its specific deliverables. When the final iteration is accepted, you can close the entire project. You'll determine whether the traditional or agile approach makes sense during project initiation, once you know whether or not your solution is clear.


## How organizational structure affects projects


From the classic functional hierarchy, where each person reports to only one supervisor, to a matrix, where people report to both functional managers and project managers, to a projectized organization, where most of the people work on projects. Each of these structures affects how projects are performed. In a functional hierarchy, projects aren't the priority, making it difficult for projects to succeed. Project managers have almost no authority. A functional manager is typically in charge of things like the project budget. Resources are hard to come by because they report to functional managers not the project manager. Even the project manager and other project management staff have to split their attention between the project and their regular work. The second type, matrix organizations, are still functional hierarchies but they support projects more than pure hierarchies do. They can be weak, balanced or strong depending on how much emphasis they put on projects. In a matrix, project managers have some authority to make decisions. Resources assigned to projects report to two managers, their functional manager and the project manager. In a strong matrix, the project manager and project admin staff work full time on projects. Projectized organizations are all about projects. This third type makes it easier for project managers to produce results. Project managers have almost complete authority over their projects, including the budget. Resources are dedicated to project work and report to the project manager. Project managers and project admin staff also work full time on project work. Organizational structure has a big influence on how projects are performed, how much a project manager can do and how easy it is to make projects successful. Using these characteristics, determine which structure your organization uses.

## How organizational culture affects projects

Organizational culture is a set of factors that guide people's behaviors and decisions within an organization. Things like shared values, beliefs, assumptions, habits, and language. Let's talk about how organizational culture affects projects. The organization's mission and vision shape the organization's culture. Projects that support the company mission are likely to get more attention and resources. When you're faced with a tricky decision, you can use the mission to determine the best thing to do. Leadership and authority are also a big part of organizational culture. If management defines clear goals and then delegates responsibility to employees, that approach will also work in your projects. On the other hand, if authority isn't handed out often, you need to work with management to get things done and build their trust in you at the same time. Another aspect of culture is the organization's work environment. With a positive environment, people are motivated to get things done. And gathering lessons learned is relatively easy, because employees are used to providing input and striving to improve. In a negative environment, you're probably going to have to spend a lot of time managing your team. Some cultures believe in following the rules no matter what. Other cultures nurture innovation, expecting employees to try new approaches, question what's been done before, and come up with better methods. You don't have to follow the rules in a rules-based culture, but if you're thinking about breaking rules, it's important to know which ones you can break. And also think about what you'll do if your non-standard approach doesn't work. Does your organization put results ahead of procedures or vice versa? That is, is it better to follow the rules, even if you don't achieve the objective or can you do whatever it takes as long as you deliver the desired results? A project's goal is always to achieve its objectives. But it's important to know where the boundaries are in making things happen. Change management can be effected by organizational culture. If the organization is risk-averse, change management might include multiple rounds of review and require approval from several people. On the other hand, if change is viewed simply as life in the project management world, change management might be a lot simpler. With people working in locations around the country or the world, you also need to consider the cultures of your team members. People may react differently to situations or communicate differently based on the norms of their cultures. For example, in some cultures, people are taught not to show weakness, and that behavior could be interpreted as arrogance by people from other cultures. Culture has a strong influence on how things happen within projects and how decisions are made. To increase your success, you need to manage your projects taking cultural factors into account.

## Project management software options

Software tools can make your job as a project manager a little easier. Let's talk about types of software and how you can put it to use managing projects. Scheduling software comes in a variety of shapes and sizes. For most projects, you'll want scheduling software to help you build and manage your project schedule. There are lots of tool out there such as Microsoft Project, Oracle Primavera, LiquidPlanner, Smartsheet, Wrike, Asana, and more. Most scheduling tools these days work in the cloud to make collaboration easier. When you think about all the project management documents you produce, it's no wonder that a word processing program is essential. Because project documents are laid out similarly from project to project, you can build document templates. That way you don't have to start from scratch every time. A spreadsheet program is another must have for calculations and analysis. Besides being handy for budgets and other financial calculations, a spreadsheet can help analyze project risks and prioritize which ones you should keep an eye on. A presentation program like PowerPoint or Prezi, is useful for communicating project information at a high level. Or when you want to include information from other types of documents. Because a team of people work on a project you need some kind of tool for collaborating with others. Basecamp, Asana, and Microsoft SharePoint, are a few cloud-based collaboration tools you can use to share files with others, keep track of issues, or even manage your workflow. If you work in an organization that runs many large projects at the same time you should consider enterprise project management software. Enterprise level software provides tools that allow you to find resources with the skills you need. And see which resources are available when you need them. It helps you track risks, issues, and other information and even build document libraries, so team members can easily find information they need. We've briefly touched on some of the software options available. If you're researching software to use consider the following in your decisions. Your organizations culture and work environment, software budget, the number of projects you manage, and their complexity.

# First Things

## Initiate a project 

Initiating a project is about getting commitment to move forward. In other words, the customer or sponsor says yes to the project and gives the go ahead to start planning. Typically, the first step in project initiation is to assign the project manager who guides the project through the rest of the initiation process. Sometimes the project manager is assigned after the project's been approved. If you find yourself in this situation, be sure to review what was done during initiation and complete any activities that were skipped or left unfinished. The customer or sponsor should make an informed decision whether to approve the project. That's why initiation boils down to defining the project. You identify the problem the project is supposed to solve and gather information about the project, objectives, requirements, deliverables, and more. Once you have an initial project definition, it's time to prepare the project charter. This document formally authorizes the project and describes the authority of the project manager. Throughout this chapter I'll describe the key elements of a project definition and what goes into a project charter.


## Identify project stakeholders

As project manager, you need to know who has a stake in the outcome of your project. They're called stakeholders. They include the customer, project sponsor, departments involved with the project, and people who work on project tasks. You need to know what they expect from the project and how they contribute to it. It's vital to understand stakeholders' importance, influence, and interest in the project. That way, you can build relationships with influential stakeholders and make sure they're satisfied with project results. Let's start by identifying major stakeholder roles. The project customer is the person or group with a problem to solve. For the hospital scheduling project, the COO oversees scheduling, so she is the customer. The project customer brings three crucial things to a project. First, the customer funds the project. In the hospital, the CEO oversees the overall budget for the new cancer wing and operational improvements. But the hospital's COO is in charge of the budget for the scheduling project. Second, the customer has a lot to say about what they project will do. Third, the customer approves deliverables from start to finish. The next stakeholder role is project sponsor. The sponsor is someone who wants to see the project succeed and has enough formal authority to make that happen, like an executive who believes in the project. For the scheduling project, the COO is also the project sponsor. She wants to see it succeed because she believes effective scheduling not only improves patient care and medical results but also produces better financial results. Her position gives her enough authority to champion the project. A sponsor can help prioritize objectives, talk to stakeholders who aren't being supportive, and suggest improvements to the project plan. The third type of stakeholder is a functional or a line manager. Functional managers run departments and are accountable for achieving their department's goals. They also manage the people in their departments who are the ones you need to staff your project. Team members are also stakeholders. While they're assigned to your project, their jobs depend on their assignments and may depend on how well they perform. Finally, there are departments or people who affect the project, and there are departments or people who are affected by it. Both of these groups are project stakeholders. For example, the project will affect how hospital facilities are scheduled, so the facilities department is a stakeholder in the scheduling project. As a project manager, you need to know a lot about stakeholders in order to keep them happy. The first step is knowing who they are.

## Analyze project stakeholders

It can be challenging to figure out who the stakeholders are for your project, which ones are most important, and the best way to work with them. That's where the stakeholder analysis document comes in handy. You can store information there as you identify stakeholders and learn about their part in your project. Identifying stakeholders and their part in the project doesn't happen all at once. You'll figure all that out over the course of defining the project. First, you need to know how your stakeholders are connected to your project and what motivates them. Start by including the department, business unit, or company the stakeholder belongs to and their position in the organization. Next, determine who the stakeholder listens to. That can help if you need to figure out how to deal with that stakeholder. You can ask that person how to handle the stakeholder, or ask them to help directly. Identify the project objectives that the stakeholder cares about and how they prioritize those objectives. That way, you know which stakeholders to talk to if an issue comes up with an objective. Categorize each stakeholder by their influence and interest in the project. That helps you prioritize stakeholders so you can manage the time you devote to working with them. Finally, document the stakeholder's contributions to the project, so you know what to expect from them, or who to turn to for things you need. For practice, use the stakeholder analysis spreadsheet included with this course. Fill in information about the stakeholders for one of your own projects. In addition, the exercise files include a filled-out stakeholder analysis spreadsheet for the hospital scheduling project. Keeping stakeholders happy can be overwhelming and time-consuming if you don't know what they want and how they're involved in your project. That's why analyzing your project stakeholders is a key task during project initiation.

## Identify the project goal

The first step toward project success is finding out what the customer really wants. The project goal is the end result that the project wants the customer to deliver. A goal is something like a problem to solve, or an opportunity to take advantage of. It drives everything that happens in the project, so you want to make sure you've got it right. You start by putting together a problem statement that clearly defines the problem or opportunity. It doesn't have to be a long-winded affair. If you can fit in to on sentence, all the better. Developing a problem statement can be challenging, because people often jump straight to solutions. Solutions describe the end result, not the initial motivation. 

## Define project objectives

Once you define the project goal, you can flesh it out by identifying more detailed objectives. They help define the project scope, the approach you choose, and the success criteria you have to meet. Business objectives support your organization's goals. For example, increase market share to 25% or provide world-class health care. Financial objectives are all about money. Think increasing revenue by 15% or cutting costs by 10%. Quality objectives specify how good results need to be. The hospital might want to decrease staff infections by 80% or decrease readmittance by 75%. Projects can also have technical objectives. These are similar to technical specifications for equipment. The hospital might want mobile equipment that can be used in the hospital, an ambulance, or at a remote emergency site. There's also a catch-all category called performance. Your project might need to finish by a specific date. The hospital has to use the government grants before they expire. Now let's talk about best practices for documenting objectives. One way to define clear objectives is to use SMART criteria. Specific objectives are clear so there's no misunderstanding about what you're supposed to achieve. For example, increase revenue by 15%. With vague objectives, you probably won't get what the project customer had in mind. Measurable objectives make it easy to see whether or not they have been achieved. For example, the hospital can measure the decrease in readmittance or revenue increase. With qualitative measures, you might use survey responses such as a minimum patient satisfaction rating of 4.5 stars. Achievable or realistic objectives spell out what you can do with the resources available. Challenging objectives can motivate people but your team might give up if they think the objectives are impossible. Time-related objectives identify when they need to be achieved. Be sure to set a clear target date. For example, the government grants must be used by December 31st, 2020. Now that the project goal and objectives are identified, you work with the appropriate stakeholders to perform a benefit analysis. 

## Choose a strategy

You're likely to find that there's more than one way to achieve your project goal and objectives. Let's look at how you can evaluate alternative strategies and select the right solution. First, bring together a small group of people familiar with the project to brainstorm strategies. As a group, read the problem statement, goal, and objectives, and then begin generating possible strategies. Brainstorming should be a free flow of ideas. The point is to get as many ideas written down as possible before you begin evaluating them. Once you've identified possible strategies, you evaluate them. A decision matrix helps you compare the options. First, the group should evaluate how well each strategy satisfies the project objectives. One way to quickly shorten the list of contenders is to check whether a strategy satisfies all the must-have objectives. If a strategy doesn't satisfy one of your must-haves, you don't have to fill out the rest of the values for that strategy, like option four in this example. Next, rate the performance for each objective. When some objectives are more important than others, increase their weighting. The strategy with the highest overall rating is most likely your winner. For the top strategy from the decision matrix, the group then asks, "Is this strategy feasible?" For example, strategies that use new technology or unproven methods might not work. If feasibility could be an issue, a feasibility study can explore whether the strategy will work without committing too much time or money. Next, consider whether the strategy's risks are acceptable. You won't know all the risks at this point. You can perform an initial risk analysis to see whether any risks are so significant and likely that a different strategy makes more sense. Finally, consider whether the strategy fits the culture of the organization. Trying to force a strategy that doesn't fit with the culture is a losing battle. Believe me, you won't get the commitment you need from management or team members. Evaluating alternative strategies 

## Gather requirements

Now that you have the project goal, objectives and strategy identified, it's time to describe specifically what the project must deliver, known as requirements. Getting requirements right is crucial. If you don't identify a project's true requirements, the stakeholders won't be satisfied with the project results. And if you include requirements that aren't necessary, the project will probably take longer and cost more than it should. For the hospital scheduling project, one objective is rescheduling procedures decreased by 75%. One requirement for the scheduling system might be simultaneously schedule staff, equipment, and facilities required for a procedure. Another requirement could be search for a next time slot that all selected items are available. These requirements are both necessary and specific. You can expand them to provide even more detail. Gathering requirements can be challenging. Stakeholders might describe their requirements incorrectly or provide inconsistent or contradictory requirements. They might leave out requirements they need or include nice to haves. Sometimes people who aren't stakeholders try to squeeze their requirements into your project. In addition, stakeholders often balk at committing the time it takes to define requirements. There are several techniques for gathering requirements. Interviews are one approach. The key is to interview the right people and come with a list of questions. If your project involves several groups, try brainstorming sessions or focus groups with representatives from each group to discuss the requirements for the project. These meetings may help you obtain buy-in from the departments that attend. Another approach is to observe how people work. In other words, watch what people do in their day-to-day activities. To validate the requirements, write them up, and review them with the workers. Questionnaires and surveys are another way to get requirements from stakeholders. Build these documents carefully so the questions don't influence the answers people provide. If documentation or results already exist, 


## Identify project deliverables and success criteria

Put simply, project deliverables are the results that a project delivers. To determine whether deliverables are what they're supposed to be, you need some way to measure them. Those measurements are called success criteria. Deliverables can be tangible. Like a building, a new product, or a new service. They can also be more abstract. Like improved results. Say, a decrease in reported errors. Deliverables help you define the project scope, which basically means what is and isn't included in the project. Deliverables then help you measure progress while your project is underway. To document project deliverables, start by listing the end deliverables. That is, the results your project delivers at the end of the project. For example, one end deliverable for the scheduling project would be new scheduling system and processes launched. Next, document intermediate deliverables, which are delivered during the course of the project. An intermediate deliverable for the scheduling project would be to sign a contract with the scheduling system vendor. Keep in mind the project customer doesn't necessarily receive intermediate deliverables. For example, you might have an intermediate deliverable for completion of initial system testing that the customer doesn't have to approve. Try to define deliverables that can be accomplished in the timeframe between status reports. That way you can evaluate progress based on the deliverables completed since the last report. Now that you've identified your deliverables, how can you tell if the ones you receive are what the stakeholders want? You guessed it. You need quantifiable criteria you can measure them with. Success criteria are definitions of what success looks like. Some are easy to figure out, like signed vendor contracts, or the certificate of occupancy for a building. Other criteria are not so easy, because they're subjective. To be effective, write success criteria 

## Identify project assumptions and risks

Another aspect of defining a project is to identify project assumptions and risks. Let's start with assumptions. During project initiation, there's a good chance you won't have all the information you really need. Don't worry. You can make assumptions about that info so you can move ahead with the project. Later when the project picture becomes clearer, you can revisit your assumptions and modify them if necessary. For example, with a scheduling project, you won't know how much system customization is needed until you choose the vendor, and that could affect how many in-house resources you'll need, which could affect how long the project takes. You get the idea. You might assume that the system won't require much customization. From that, you estimate that you'll need three in-house IT resources, and they'll need a month to finish their part of the project. Assumptions can be tricky when people don't realize they're assuming something. In that case, they assume something is true without confirming it. Different people can make different assumptions. If those assumptions aren't brought into the open, someone will end up disappointed. For example, say that you assume that the system vendor is going to load data into the database and the system vendor assumes your internal IT group will do that task. If that assumption isn't identified, the system might miss an important deadline. The key is to get assumptions out in the open. That way you can make sure everyone is on the same page. Think about uncovering unspoken assumptions as you define and plan the project. Ask people what they expect and what they envision when they think about the project. Don't be shy about asking more than once. Now let's look at risks. A risk is a situation or event that might occur that could affect your project positively or negatively. What makes a risk a risk is that it's uncertain. 

## Prepare a project scope statement

After all the work you put into defining your project, it's time to get that project definition in writing. Project scope is the culmination of your definition work. It describes the boundaries of the project, what is included in the scope of the project, and just as important, what isn't included. It's important to get the project scope in writing. Otherwise, you're bound to run into scope creep. That isn't some weird guy who casts furtive glances at your project scope. Scope creep is when stakeholders ask questions like: Can you do this one thing? We forgot to ask for this. Can you add it to the project? And you say yes without running the change through your change management process. The other reason you want project scope in writing is to remind stakeholders what they agreed to at the beginning of the project. If someone says, "I thought you were going to do X, Y, and Z," you can go back to the scope statement and show them that those items are out of scope. If they really want those things in the project, you can use change management to add them. A scope statement is a document that spells out the project scope. Everything you've done during project initiation feeds into this document, the goal and objectives, deliverables and success criteria, assumptions, risks, and constraints. Here's what is within scope for the hospital scheduling project. Your scope statement also includes an out of scope section that shows what the project doesn't include. For example, the scheduling project doesn't include an update to the system used to assign staff to work shifts. Scheduling resident rooms in the hospital and rehabilitation wing is also out of scope. Keeping scope under control is one key to a successful project. Use the scope statement template in the exercise files folder to create a scope statement for the hospital scheduling project.

## Create a project charter

Is the project definition done? Check. Is the scope statement ready? Check. Now you're ready to get approval to move into planning your project. You also put together a project charter that officially authorizes the project and publicizes it to the world. The whole point of defining a project is to give the project customer or management team the info they need to approve the project. The review process varies from organization to organization. Typically the customer compares the project information to acceptance criteria that align with the organization's goals. There are three possible outcomes from this review. The project is approved to proceed to planning, it's denied, or it's sent back for rework. When the project is approved the final step in initiation is to create and distribute a project charter. This document authorizes and publicizes the project. Here's what typically goes into a project charter, the name of the project, the purpose of the project, a short summary of it's goal and objectives will do, a high-level project description, which may include things like high-level success criteria, requirements, scope, risks, assumptions, and constraints, a milestone schedule, cost estimate, and list of stakeholders. Remember, you'll flesh out all of these components once you start planning your project. You also include information about the project manager, the projects manager's name, the project manager's responsibilities, including a brief description of the work the project manager does. The extent of the project manager's authority, and the specific work the project manager has authority to perform, such as requesting resources or signing contracts. Finally, the charter includes a formal declaration of the sponsor's support for the project. Think of this declaration as a power of attorney given to the project manger by the sponsor or customer. You might wonder why a project charter describes the project manager's authority. It's because project managers don't have the kind of authority that managers in a structured organization do. Project managers authority lasts as long as the projects they manage, and applies only to those projects. For that reason it's important that people understand what a project manger is authorized to do. When the project charter is ready to go the project sponsor distributes it to everyone affected by, or involved in the project. Once the project has been authorized and your authority as project manager is common knowledge, you're ready to begin planning the project.

# Develop a Project Plan

## What is a work breakdown structure?

After you get the go ahead to plan your project, the first thing you do is identify the work that has to be done. Even small projects can have lots of tasks. That's why project managers use a diagram called a Work Breakdown Structure, or WBS to organize work. You use it to divvy up project work into manageable pieces so you can plan, track, and manage your project. Creating a WBS helps manage the project in several ways. First, it's easier to estimate time and cost for smaller chunks of work. Say the sponsor wants to know how long the whole project will take. You can't come up with the good answer off the top of your head. It's a lot easier to think about how long smaller parts of the project will take. After you estimate the small chunks, you can add them up to get an estimate for the entire project. It's also easier to assign work to team members. By breaking work into smaller pieces, you'll have built-in checkpoints to measure progress on the project. A WBS contains two kinds of tasks, summary tasks and work packages. Summary tasks are the higher-level tasks that summarize work in some way. Summary tasks describe parts of the project. They could represent phases or deliverables or even work done by different groups in your organization. In our scheduling project, the WBS is organized by deliverable. How many levels of summary tasks depends on the size of the project. A few levels are enough for a small project. If a project is large or complex, you might have several levels of summary tasks. Work packages are the lowest level tasks in the Work Breakdown Structure. They spell out lowest level project deliverables and the detailed work needed to deliver them. In the scheduling project, the summary task to configure the system server might have work packages to activate modules, set options, and add software links. Breaking your project down into manageable pieces helps you plan and manage your project effectively.


## Build a work breakdown structure

The best way to build a work breakdown structure, or BWS, is to start at the top level of summary tasks and work your way down. Work with your team to identify the WBS. You're less likely to forget work that needs to be done, and the team is more likely to buy into the project. As a team, you identify the top few levels of summary tasks. Then you can have smaller groups break down the summary tasks into smaller chunks. At the end, everyone gets back together to review the results and correct any issues. If the full team isn't onboard yet, work with your initial team to brainstorm the WBS. Then you can revise and add more detail to the WBS later as others join the project. Start by using the scope statement and deliverables to identify the top level summary tasks. For example, the scope for the scheduling project includes redesign scheduling processes, a new scheduling system, documentation, and training. So you add summary tasks to cover all of those. Next, break down each of the summary tasks into smaller pieces. Intermediate deliverables come in handy for identifying lower level summary tasks and work packages. For example, the scheduling project includes a deliverable for a signed contract with the system vendor. Let's add tasks to obtain proposals, select the vendor, negotiate terms, and sign the contract. You might wonder, how much break down is enough? Most project managers shoot for work packages that take between eight and 80 hours to complete. Work packages are the lowest level tasks in the WBS and represent the work that needs to be done. Consider breaking down work to match the frequency of your status reports. That way, you have measurable progress and completed tasks for every status report. Here's some tests to determine whether you've broken work down to the right level. Time and cost are easy to estimate. Status is easy to measure. Task durations are shorter than your reporting periods. The detail is at the level that you can and want to manage. Different parts of a project might require different levels of decomposition. One part of the project might include more work, so you break it down to three or four levels. If another part of the project is simpler, you might need only two levels. Don't worry about the initial organization of your WBS. You can rearrange it later as you learn more about the project. Building a WBS from the top down works well because you can use the scope statement and deliverables to get started. For practice, try building more of the scheduling project's WBS based on the project scope and deliverables.

## How to create work packages

A short task name in a WBS isn't enough to tell team members what they're supposed to do. To give your team the info they need, create work package documents that describe the work that needs to be done in detail. The level of detail needed in the work package document depends on things how familiar the work is and the experience of the person assigned to the task. For example, a work package document might include a detailed description of the work for a less experienced person. For someone more experienced, a checklist of things to do might be enough. If the specifics of the work are described somewhere else, you can refer to the other document that contains the information. A work package document does more than describe the work, it also identifies how you know the task is complete and whether it was completed correctly. For some tasks, you can include the corresponding deliverable and success criteria in a work package document. Otherwise, write up a description of what you will have when the task is complete and what it should look like. As project manager, you won't know enough about every aspect of the work to produce these detailed task descriptions. Turn to the people who helped you build the WBS, team leaders for the groups that will work on the project or other knowledgeable people to help fill in the details. Work packages help your team members deliver what they're supposed to. For practice, try defining a work package for the schedule training task that appears in the scheduling system WBS.

## Estimate time and cost

The two questions you get hit with most often about a project are: "How much time will it take?" And "How much will it cost?" Estimating accurately matters because your estimates could determine whether it makes sense to run the project. You start by estimating time because it affects both project schedule and cost. You also estimate cost that aren't time-based, like materials and ancillary costs like travel. During initiating and planning, you might work with a core planning team to develop initial estimates. During project execution, you can get more accurate estimates from the people assigned to tasks. They understand what has to be done and know how long it wil take based on their experience. Plus, they're usually motivated to live up to the estimates they provide. Estimates don't have to be perfect right off the bat. For project selection, an estimate that's plus or minus 75% might be good enough. As you learn more about the project during planning, your estimates grow more accurate. Ideally, to plus or minus 10%. There's several techniques you can use for estimating. If you have similar projects that are already complete, use them as a foundation for your project estimate. With parametric models, you calculate work and cost based on some measure, like the number of square feet for construction. This technique works when you have data from many, similar projects. If a project represents uncharted territory for your organization, consider bringing in experts who are familiar with the work, like consultants or vendors. The Delphi technique counts on several heads being better than one. First, you ask several experts to produce estimates independent of one another. You share the results with the group, keeping the estimates anonymous. You keep the estimates anonymous because you don't want anyone to be influenced by the reputation or authority of a co-expert. You then ask everyone to estimate again. Repeat this step a few more times and then use the average of the last round as your final estimated value. For large projects or rough estimates, top-down estimating works well. You estimate phases or major components and then break those estimates into smaller pieces until you get to individual tasks. Estimating from the bottom-up means you estimate each task and then add them up until you have the estimate for the entire project. You can also start with a top-down estimate and then revise it by working your way back up from the bottom to the top. The project schedule and cost hinge on your estimates, so it's important to make them as accurate as possible. For practice, based on what you know about the hospital scheduling project so far, which estimating technique would you choose and how accurate an estimate would you strive for?

## How to choose the best estimate

Project estimates are a lot like your morning commute. You need to choose an estimate that gives you a good chance of getting where you need to be when you need to be there. On some days, your commute is a breeze. Maybe 30 minutes. Most days, let's say it's 45 minutes. But an accident can turn it into an hour. For your commute you choose a drive time that gives you some wiggle room so you get to work on time. The same approach works equally well for your project estimates. You probably decide when to leave for work informally. We're going to look at a similar, yet rigorous approach to choosing estimate values for your project. If you were to graph the probability of your commute times, it would look like a Bell Curve with your average drive time at the peak. A Bell Curve is also known as a normal distribution. And it has properties that help you choose a safe estimate to give your customer. First, on a Bell Curve, 50% of the possible values are less than the average, and 50% are greater. That means the average value gives you a 50-50 chance of your results being equal to or less than your estimate. Think about it. If you give your customer your average estimate, you're as likely to fail as you are to succeed. I wouldn't stake my career on those odds. On the other hand, the worst case estimate gives you the highest chance of success. However, that number is so high your customer might cancel the project. You probably know that customers are prone to ask, "What's the best case scenario?" "What if everything goes right?" Don't fall for this trap. Suppose you give the customer your best-case estimate. Say, six months. And you describe everything that has to go right to finish the project in six months. This is what they'll hear. Blah blah blah blah six months. And you've set up your project to fail. We've eliminated the best, worst, and most likely values. You're probably wondering what number you should use for your estimate. The answer is about halfway between the average and worst-case values. Let's go back to the Bell Curve and pick the value halfway between the average and worst-case values. Without getting into the mathematical details, that gives you an 86% probability of your results being less than or equal to that number. That's a good chance of success, and the value is easy to calculate. Plus, you can tweak your value higher to reduce the chance of missing the deadline. Just like you would leave extra early if you're going to a job interview. And if you're trying to win a project, you might choose a smaller value. Even though the risk of failure is higher. The bottom line is, choose the estimate that provides an acceptable probability of success. For practice, use the information in the exercise files to determine the time estimate you would give the hospital's COO.

## Create a resource management plan

Part of planning is to identify and document project roles and responsibilities, who reports to whom, and the skills needed to do the project work. That's where the resource management plan comes into play. This plan also includes a staffing plan, which describes everything about how you'll staff the project. You start with a responsibility matrix. This document spells out who can make or approve decisions, the groups performing work, and which groups need to be consulted or informed about what's going on. A responsibility matrix includes four categories of responsibility. R means a group is responsible for performing work. I represents Informed, which means a group gets information. C means that you consult a group about decisions. However, they aren't accountable for the decision that's made. A is for Accountable, which means the group makes or approves decisions and delegates work. During planning, review the responsibility matrix with stakeholders and work out any disagreements. If part of your project doesn't have an owner, meaning someone accountable for that area, talk to the stakeholders and your project sponsor to identify who is accountable for that area. You might have outsourcing, subcontracting, and partnering arrangements. Don't forget to add these groups and how they contribute to the responsibility matrix. The second part of a resource management plan is a project organization chart. It's like a regular org chart, except it shows the hierarchy and reporting structure for people involved with the project. That way, you know who to talk to if you need to escalate a request or a decision. Third, identify the skills the project requires and how many people you need with those skills. A skills matrix helps you figure this out. To build one, take a look at your work packages and identify the skills each package requires. Then create a matrix with your project tasks in rows and the skills you need in the columns. Add checkboxes in the matrix when tasks require a specific skill. Then you estimate the number of resources you need with each skill. You can even multiply that number by the average pay rate to estimate the labor cost. Finally, it's time to develop the detailed staffing plan. Identify where you plan to get resources. Are they in-house, outsourced, or contracted? When do you need resources? Identify any training they need and document resource-related processes, getting team members on board, handing out assignments, getting status updates, and releasing them from the project when their work is done. A resource management plan documents who does what, who reports to whom, the resources you need, and how you'll manage them. As an exercise, use what you know about the project so far to complete the responsibility matrix in the exercise files. Then, using your responsibility matrix, draw up a project organization chart.

## Build a project schedule

A WBS identifies the work needed to complete a project. But it doesn't tell you how long the work will take or when it can be performed. To do that, you need to turn your list of tasks into a schedule. First, put the tasks into the right sequence. You add dependencies or links to tasks to identify which tasks have to finish before other tasks can start, which tasks finish at the same time, and so on. For example, the task Build Specs has to finish before Review Specs can start. Likewise, Review Specs has to finish before Approve Specs can start. We'll look at dependencies in detail later in the course. For now, the most common dependency type is finish-to-start. Next, you add the time you estimate each task will take. Time estimates and task sequence help determine how long the entire project will take. Third, you assign the people on your project team to tasks. With the estimated time and people assigned to do the work, you can calculate the task duration. Finally, take into account other constraints like deadlines and when resources are available. If your initial schedule doesn't work, you can tweak it in various ways, whether you're trying to shorten the schedule, cut costs or even out people's workloads. The schedule is one of the most important aspects of your project. It tells you how long the project will last and when you need the people who will do the work.

## Develop a project budget

With most projects, money is important, whether you're trying to make it, save it, or stay within a budgeted amount. Project cost is one of the three key constraints you need to manage and the project budget is your starting point. The project budget should be a realistic estimate of the cost to complete the project work. If your estimate is unrealistically high, the project could be canceled. If the estimate is too low, the project will cost more, which usually means it won't deliver the financial benefit it's supposed to. You estimate all the costs associated with completing the project work, including labor, materials, and other costs, like travel. You can start with rough estimates for the first few levels of your work breakdown structure. Then create more accurate estimates as you learn more about the project. Labor cost is usually a big part of a project's cost. That is, what you pay people who work on the project. If you hire vendors or contractors, their charges go directly into project labor costs. For employees, you typically use the burdened cost. That includes salary and employee benefits. You can get burdened rates for roles from your accounting or HR department. Projects might use other types of time-based resources. Think equipment or office space that you rent. These costs go into your calculations, too. Then there's material cost. The hospital scheduling project could include a new server or training guides and documentation notebooks. Finally, be sure to include other types of costs that don't fit in the previous three categories. For example, travel, training expenses, and fees. Have you ever gotten into a cash crunch when you didn't have enough money on hand to pay expenses? Projects can run into cash flow problems, too, which is why you figure out when project expenses will occur. When you assign resources and other costs to the tasks in your projects schedule, you'll get a rough idea what your cash flow looks like. Let's say management has money allocated for the project, like the grants and donations earmarked for the hospital scheduling project. If your cost estimate is higher than the allocated amount, you have to look at ways to reduce the project cost. You might eliminate nonessential expenses or find less expensive resources, maybe cut scope. In the hospital scheduling project, management allocated *$850,000*. The estimate, including contingency, is *$950,000*. You might consider working with less contingency funds, ask for more money, or dive in to find ways to cut the total cost. The project budget is the cost target you aim for as you manage a project. For practice, think about what you can do about the *$100,000* that exceeds the amount that management has allocated for the scheduling project.

## Identify risks

You've already put a lot of planning into your project. The next step is to plan for what could go wrong because you know something will. Instead of asking for an adrenaline rush why not identify what could go wrong, so you can plan ahead for dealing with issues. Let's start with risks you're aware of. They're called known unknowns. Things like weather delays, or resources being unavailable. Let's look at some examples of things that can introduce risk to projects. Technology, it might cost more than you expect, not work the way it's supposed to or not show up when you need it. A widely dispersed project team can increase risk. You could have delays due to time zone differences, miscommunication from different languages or cultures or lack of teamwork because remote team members haven't developed effective working relationships. Lack of detail like specifics on deliverables, due dates or who will work on your project can lead to all sorts of problems. Limited options like people with rare skills increase risk because you don't have easy alternatives to apply if a problem arises. Work with everyone on your team to identify risks. You can ask experts for different parts of the project about the risks they foresee. Ask key people on the project what they think and ask other project managers about risks from similar projects. Fill out a Risk Information Form for each risk you identify. Include as much detail as you can, which objectives are in danger? What events can cause the risk to occur? What the consequences are? Who owns the risk? And so on. Up until now, we've talked about risks we can anticipate but what do you do about the risks you can't predict? These are called unknown unknowns. They come from situations that are so unlikely that they never occur to you. You're unlikely to have to deal with such a significant unknown risk but you will encounter unknown risks at some point and you need to plan for them. How? By setting aside contingency funds and time for dealing with them. The question is, how much should you set aside? Many companies use a percentage of the project budget, like 15%. But the percentage is typically based on past experience. To minimize the effect of risks, you first have to identify the risks that your project faces. Try to identify several risks that the hospital scheduling project might face. For an extra challenge, use the Risk Information Form in the exercise files to document the risks.

## Create a risk management plan

A risk management plan documents which risks you're going to worry about and how you plan to handle them. First, identify which risks warrant managing. Start by evaluating how likely they are and how serious their impact would be. Rate risks by probability and impact using a scale of one to five, where one is unlikely or not serious, and five is highly likely or extremely serious. Then multiply the probability and impact to calculate a risk score. It makes sense to manage risks that are at least fairly likely and serious. That means you need to plan for any risk with a score of nine or more. For example, suppose older hospital computers are very likely to be incompatible with the new scheduling system, new computers increase the budget so the impact is very serious. The risk score is 25, so you'll definitely need to manage it. Risk one's probability and impact are both medium. It's score is nine and you need to plan for it. Risk four is fairly serious but it isn't likely. It's score is three, so you won't plan for it. Next, you plan how you'll handle each risk. The goal of a risk response is to plan the actions you would take to reduce adverse consequences a risk might have on your project. Here are some risk responses you can use: the easiest option is to accept the consequences. That's fine for risks with low probability and impact. Avoiding risk is another approach. For instance, changing the project scope to remove the risky part of the project. Mitigating risk means taking steps to reduce the consequences or impact of the risk. For example, by running a feasibility study, you'll have a better idea whether a scheduling system will meet the schedule, budget and other objectives. Transferring risk means handing off the risk to someone else, like when you purchase insurance, this reduces risk but doesn't eliminate it. Insurance might relieve costs, it won't eliminate other factors like time delays. Make sure your responses aren't overkill. If a schedule delay triggers a *$500* late fee, it doesn't make sense to spend *$5,000* to prevent the delay. Remember, contingency, time and funds can be used to handle risks that have no other response, or to handle risks you didn't identify upfront. The final step in a risk management plan is defining how to monitor risks and measure responses. Create a risk log to summarize the risks you'll manage. For each risk include a description of the risk, the events or circumstances that trigger the risk, the probability and impact, the response you've chosen, who will monitor the risk, the result you expect from the response if you have to use it, and finally, the risk status. Keep your risk plan up-to-date as the project progresses. Some risks become less likely as time passes, other risks may be introduced because of project changes. Risks occur so you track what caused them, their impact and whether your responses are working. A risk management plan helps protect your project from things that can go wrong. For practice, fill out more of the risk log in the exercise files.


## Set up a communication plan

If you've ever heard Abbott and Costello's "Who's On First" skit, you know how hard it can be to get your point across. The success of a project depends on good communication. That's why you set up a communication plan, the right people, getting the right information at the right time and in the right way. You start with your audiences. Who needs to know about the project? The responsibility matrix is a good place to look to find your audiences. Second, what do your audiences need or want to know? Management stakeholders care about a project achieving its objectives. Early on, you give them the project plan so they can see how you're going to deliver. Later on, they want to know about progress, how much you've spent, and the overall project result. The project sponsor is tightly connected with the project. You'll probably communicate with the sponsor more often, maybe in face to face meetings. Functional managers initially need to know the skillsets you need, when you need them, and other things like cost constraints. Once people are on assignment, these managers want to know when their people will be done. Team members need to know what they're supposed to do. They're more likely to buy into the project if you tell them how their work fits into the big picture. As time passes, they also need to know what's coming up next. The last part of a communication plan is the how: how often, the methods you use, and the format. Face to face communication is good for brainstorming, getting to know people, and discussing sensitive topics. If your team is dispersed geographically, email, conference calls, and video conferencing are indispensable. You know that stakeholder management is all about getting and keeping stakeholders' support for your project. The communication plan helps by spelling out how you're going to communicate with stakeholders to keep them onboard and happy. A communication plan helps ensure that people get the information they need in the most effective way. To solidify your understanding, set up a communication plan for the hospital scheduling project.

## Develop a quality plan


Project success actually means meeting objectives, not exceeding them. Quality management includes processes to make sure that a project satisfies it's objectives and requirements. A quality management plan is your roadmap to meeting objectives. Project quality translates into meeting the customers requirements and delivering on time and within budget. With deliverables in products, quality also means conforming to specifications. A quality management plan has three components. First, there's the quality standards for deliverables. For a product the quality standard might include the acceptable tolerance for it's dimensions, or the acceptable number of defective widgets in a batch. For the hospital scheduling project, one quality standard might be that the scheduling system meets all the required scheduling functions. Keep in mind that the goal is to meet the quality standards. Obviously, you don't want quality that's below standard. The customer won't be happy. They might ask for additional work to get it right, or simply refuse to pay. This might surprise you, you don't want quality that's higher than required. Why, because other aspects of the project can suffer like a later finish date, or higher cost. The second part of the quality management plan is a quality assurance plan. This spells out the processes you use during the project to ensure that the final results meet the quality standards. For the scheduling system, reviewing requirements and designs falls into this category. You might also test software functions as they're developed. Finally, you plan for quality control which means how you measure and monitor quality in the final deliverables. The bottom line is that you examine or measure results to see if they meet the standard set. For the scheduling project the final acceptance test would measure whether the processes and system achieve their objectives. For the hospital's new cancer wing you would do a walkthrough to confirm that the construction has been completed correctly. Continuous improvement is a big part of quality management. If you find quality issues, you analyze the problems to see how to prevent them, or at least reduce their frequency. Cause and effect diagrams are also called fishbone diagrams, because they look like a fish skeleton. They help identify factors that could lead to problems. Those factors can help you figure out ways to prevent the problems. Another tool is a Pareto diagram. This shows how many defects are generated by each cause. That way you can address the causes that produce the most defects first. By developing a quality management plan you can ensure that your project meets it's quality objectives. For practice, identify potential quality assurance processes, and quality control measures for each of the project objectives.

## How to set up a change management plan

You know changes are going to happen, so the only solution is to manage them. A change management plan helps you add important changes into your project while keeping out the ones that don't make sense. First you identify what you want to control like the project scope, requirements, and the schedule. Or perhaps the entire project plan. The versions you control are called baseline documents. The list of project requirements approved by the stakeholders is your baseline. If new requirements are suggested, you use your change management process to decide whether or not to add them to the requirements list. You also need a group that reviews change requests and decides whether to approve them. It's called the change review board and it's usually made up of key stakeholders. Then you define a change management process. That's what will happen when someone requests a change. The process your organization chooses depends on things like the company culture and project size. Here are the typical components of change management processes. The first process is documenting and submitting a change request. Use a standardized change request form that requesters fill out. It's easier to evaluate changes because each request has the information the review board needs to make decisions. Include details about the requested change, the reason it's needed, the business justification, and the results it should produce. Next, someone has to evaluate the request and estimate its impact. As project manager, you can assign this task to someone on your team. The evaluator determines whether the change is needed. If it is, the person evaluates whether the suggested approach is correct or if an alternative would be better. The evaluator estimates the effort and cost the change would require, the impact on the project, and whether it introduces risks. Third, the change review board reviews evaluated change requests. The board might accept or reject the change request. They also might ask for more detail or revisions. Be sure to communicate the decision to the person who requested the change. For approved requests, you update the baseline documents to reflect the change. For example, your project schedule might need new tasks or deadlines extended. Finally, track where change requests are in your process. A change request log shows the status of submitted change requests, who's in charge of the request, the impact estimate, current status, and at the end, actual impact. Consider setting thresholds so team leads can decide what to do with smaller requests. That way, your change management process isn't overwhelmed with small items. Also, create a process for emergency changes that need a rapid decision between meetings of the change review board. A change management plan helps you include changes that make sense in your project while protecting your project from being overwhelmed with unnecessary changes. To solidify your understanding of change management, draw a flow chart of the change management processes the hospital might use.

## How to plan procurement

Even with homemade chicken soup you don't make everything from scratch. Imagine how long it would take and how much it would cost if you had to raise your own chickens, vegetable and wheat and make chicken stock and noodles. The same goes in the project world, procuring products services and skills can be a lot quicker than trying to do everything yourself. When you purchase items from outside your organization, you need to plan how you're going to do that. You guessed it! That's what a procurement plan is for. First you identify what you need to purchase. Maybe you need people with specific skills or you need more people than your organization has. Or the project requires products and materials. Second, you document your procurement processes. Who handles procurement? Is it the purchasing department, the project or team or a combination of the two? The plan describes criteria for choosing vendors, the selection process, types of contracts you use and how you manage those contracts. Fortunately, you usually don't have to invent all these processes. Most organizations already have procurement processes in place. Talk to your purchasing department or finance group to find out how it's done. Third, you describe your make or buy decision process. Basically, how you you decide whether to use in-house products or services, or procure them. To make an informed make or buy decision, it's important to understand your needs. Make sure your requirements are clear and then prioritize them. That way it's easier to pick products that meet your requirements, plus you can resist bells and whistles you don't need. You have to see if a product that meets your needs is available. If it is, evaluate how it matches up with your requirements. Before you finalize your decision, consider the pros and cons of making versus buying. If your project is on a tight timeline, buying is probably the right choice. The last part of the procurement plan is a list of potential vendors who offer what you need, to purchase. Describe how you researched vendors and contractors and the criteria for developing the list. Procurement planning helps you make smart decisions about whether to purchase products and services, and if you do, which products and services to procure. As an exercise, identify some of the criteria you might use to choose the scheduling system vendor.

## Processes for procuring resources

For initiating and planning, you might work with a small core team. Before you begin executing project work, you bring on the rest of the team and acquire needed equipment and materials. If team members come from within your organization, procurement might be as simple as telling your in-house team it's time to start your assignments. For resources outside your organization, the procurement process has a few more steps: solicitation, evaluation, selection, contracting, and ongoing management. These steps can take time so be sure to leave room in your schedule for performing them. Solicitation often begins with a request for proposal, abbreviated as RFP. It's a typical way to ask vendors for proposals or bids. An RFP describes the services or resources you need, when you need them and your budget for the work. Include the criteria you plan to use to select a vendor so companies can decide whether or not to bid. Don't forget to include instructions and the deadline for submitting a proposal, and the date you'll announce your decision. Once you have all the vendor's proposals or bids, you hunker down and evaluate the responses using your criteria. How you select the winner depends on the size and complexity of the project. Evaluation and selection might be as easy as filling in a spreadsheet with ratings and choosing the vendor with the highest score. For large jobs, you might whittle the submissions down to a shortlist of vendors and ask each one in the second round to prepare a more detailed presentation, or demonstrate what they have to offer. For example, in the scheduling system project, you might ask the top two candidates to demonstrate their systems and show how features satisfy project requirements. Contracting is when you prepare and sign a contract for the resources you need. Contracts usually include a statement of work, terms and conditions, deliverables, deadlines, and of course, the price. Contracts fall into several categories: fixed-price contracts work when your project is well defined. This type of contract places the risk on the vendor because the vendor is paid a fixed amount no matter how much time the work takes, or how much it costs. Time and materials contracts pay for time worked and expenses incurred. The project accepts most of the risk, so it's common to include and not to exceed value to limit the price. A cost-plus contract is like time and materials with the addition of penalties or rewards based on the vendor's performance. With a retainer, you agree to pay for a specified amount of time and define the work as you go. Complete procurement steps early enough so the resources you need are available when work begins. For example, if your project requires materials with long lead times, you'll need to start procurement early. After you award contracts, you don't sit back and relax. While project work is underway, you manage those contracts, monitor the vendor performance and manage any changes. For practice, think about which type of contract you would use for the scheduling system and why.

## How to obtain approval to proceed

Now that you have a plan for the project, you need to get the stakeholders to approve the plan so you can start work. Getting approval is important because you need stakeholders' commitment to make the project a success. You might think about getting approval by mailing or emailing a packet to stakeholders and asking them to sign on the dotted line. I don't recommend that approach, here's the problem. People probably won't read through the plan. Sure they might sign their names, but later if they find something they disagree with, their commitment disappears and you end up back to planning. A face-to-face sign-off meeting is more effective. Schedule a meeting specifically for getting approval. The agenda is straight forward. First, present the project plan to make sure that the stakeholders agree with it. If issues arise, deal with them right then and there. Or, if the issues are big enough, you might have to reschedule the sign-off so you can rework the plan. Once everyone agrees, ask them all to sign a signature page to indicate their approval. If you can't get everyone in the same room, a video conference or conference call works, too. Ask the people in other locations to transmit their signatures to you by fax, email or even snail mail. Signing a project plan isn't like signing a legal document. You're never going to go back to your stakeholders and say, "But you signed this." What's important is that the stakeholders understand what the project is about and buy into it.

# Build a Project Schedule

## Put tasks in sequence

A big part of building a schedule is getting tasks in the right order. Mealtime goes a lot better when you cook dinner before you eat it. By putting tasks in order, you turn your WBS into a sequence that defines when project work should occur. In the project management world, network diagram is the name for the diagram that shows the sequence of tasks. Each task appears in a box with the task name, and perhaps other task info. Arrows drawn between boxes shows how the tasks are linked. Even if you build the sequence by putting sticky notes up on a whiteboard, you're creating a network diagram. The links between tasks aren't just about which task starts first. A task dependency is when one task controls the timing of another. The task in control is called a predecessor, and the one being controlled is the successor. Each task has a start and finish, so there are four types of task dependencies. Finish-to-start dependencies are the most common. The finish of one task controls when the other task starts. For example, you have to analyze the current scheduling processes before you can start designing new ones. A finish-to-finish dependency means the finish of one task controls the finish of the other. For example, the dev team finishing the scheduling features controls when they finish testing the features, with a small delay to test those last bits of customization. Start-to-start means that the start of one activity triggers the start of the other. This dependency type can cause trouble if the predecessor takes longer than it's supposed to, as shown here. If writing takes longer, the reviewing task looks like it finishes before writing does. That would mean either some writing isn't reviewed, or the reviewers have to wait until the writing is done. In this example, the correct logic is actually finish-to-finish. With a finish-to-finish dependency, if the writing finish date is delayed, so is the finish date for reviewing. An example of a true start-to-start dependency is pouring concrete and troweling it. Because concrete hardens with time, the start of pouring concrete controls when troweling starts. Start-to-finish dependencies don't occur very often. That's a good thing, because they can be confusing. The start of one task triggers the finish of another. So the task in control occurs after the one it controls. Consider the shifts at a retail store. The first shift can't finish until the clerk for the second shift shows up. You can figure out which type of dependency to use by asking a few questions. Which task controls the other? That tells you which task is the predecessor. Does the start or finish date of the first task control the second task? That identifies whether the dependency begins with start or finish. Does the predecessor control the start or finish of the successor? That identifies whether the second half of the dependency is start or finish. Adding task dependencies to your tasks in a network diagram gets your project tasks into sequence. Now that you know all the dependency types, build a network diagram for the tasks in the partial WBS in the exercise files.

## How to assign resources to tasks

Your project tasks are in sequence and you've estimated their durations. The final piece of your scheduling puzzle is how many people work on each task and when they're available. With that info, you can see how long tasks take and when they should occur. You assign resources only to work packages, the lowest levels in your WBS, because they represent work that must be done. Summary tasks roll up work from lower levels, so you don't assign resources to these tasks. Milestones don't have duration, so it doesn't make sense to assign resources to them either. What happens when you assign resources to work packages depends on whether you estimate task duration or work. Let's say you estimate duration and assign the number of resources you expect to get. Your scheduling tool will calculate the work those resources can perform in that duration. Let's say you're expecting two vendor resources full time to design scheduling features. The task duration estimate is two weeks. Two people working full time for two weeks is 160 hours. On the other hand, if you estimate work, your scheduling tool uses the work estimate and the resource allocation to calculate the task duration. Some tasks don't get shorter no matter how many people you assign. Meetings are the classic example. A four-hour meeting is four hours long whether three people attend or 10. When resources are available also affects when work occurs. For example, your schedule might show the equipment installation starting on November 9. If your IT folks are wrapping up another installation and the team won't be available until November 18th, guess what? You have to delay that task in your schedule to when the resources are available. Don't forget to assign other types of resources you need like materials, equipment and ancillary costs. With all resources assigned, you finally get a good idea what the schedule looks like. To practice, work through the assignment exercise in the exercise files.

## Learn to use milestones 

Milestones get their name from the past when people placed stones by the side of a road to mark each mile. In projects, milestones do a similar job, but they show progress and other key project points rather than distance. They show when you've completed key tasks or portions of a project. And they make it easy to see how much you've completed and when it finished. Milestones are great as the first and last tasks in your project schedule. By starting a project with a milestone, you can reschedule the project's start date just by moving the starting milestone to a later date. By looking at the final milestone, you can tell whether the project is on time, late, or ahead of schedule. Milestones also help highlight progress you've made in between the project's start and finish. When all the work leading up to that milestone is done, you have the satisfaction of marking off another milestone as complete. If you're waiting for someone to deliver something, add a milestone to flag that delivery. Finally, use a milestone to flag decisions that determine what happens next in a project. Or use a milestone for an approval that determines when the work after the approval can start. Milestones help indicate progress in a project. When you link them to work tasks, you can also use them to reschedule portions of a project. For practice, add milestones to the Network Diagram in the Exercise Files.

## Make a realistic schedule

It's important to make your project schedule realistic so tasks occur as close as possible to the dates you planned. One way to build reality into your schedule is to estimate task duration based on the hours people work on the project each day. People don't work 100% of their time on project tasks. People don't even work 100% of every workday. Things like staff meetings, paperwork, training, and time off chew up work hours. Even walking across campus uses up work time. So do your best to estimate the number of hours people typically work. Another dose of reality is adjusting estimated task hours based on how fast the assigned workers are. For example, if someone is a whiz at getting computers connected to the network, you can change the task work hours to reflect her productivity. Third, to keep people productive, assign resources to work on no more than three tasks at a time. Switching between tasks means a person has to switch focus, and that introduces a small delay each time. When resources are in demand and you have to assign them to several tasks, adjust those assignments to reflect their decreased productivity. When you make adjustments like these, add a note to your schedule to document the change and the reason why. That way you'll know what to do if other changes come up in the future. The closer you model your resource assignments to reality, the easier it will be to keep your project on time. Complete the assignment exercise in the exercise files to practice assigning resources to tasks.

## Understand the critical path

The critical path is the longest sequence of tasks in your schedule. Why is the critical path so critical? Because any delay on that path delays the finish date of the project. Just as important, if you can figure out how to shorten the critical path, you can shorten the project schedule. How do you tell which tasks are on the critical path? Simple. They don't have any slack, also known as float. Just like a string with no slack, critical tasks can't move without affecting the project finish date. In this simple example, the System and Training tasks occur one after the other with no slack. If any of these tasks are delayed, the project finish date moves later in time. On the other hand, if a task has slack like the Reports task here, it can start later without delaying the tasks that come after it. The Reports task could delay by a few weeks before it delays the project finish. Let's look at how you tell whether or not a task has slack. A task has two sets of start and finish dates. The early start and early finish are the earliest possible dates the task could start or finish based on its dependencies with other tasks. You calculate these with what's called a forward pass. That's where you start at the project start date and use task durations and dependencies to calculate when they finish. For example, the Reports task can start as early as October 26 and finish on November 27th. The late start and late finish are the latest possible dates without delaying tasks that follow. You calculate these dates by working backwards from the end of the project. You got it, the backward pass. The late finish for the Reports task is December 18th, so working backwards, the late start is November 16th. The task has several weeks of slack, so it isn't on the critical path. Critical tasks have no slack. In other words, the early and late dates are the same. That's why the Training task is on the critical path. The good news, you don't have to perform these calculations. When you use project scheduling programs, they calculate the critical paths for you. The critical path is the place to look when you want to keep your project on time or deliver it early.  

## Understand the critical chain method

Critical chain is a slightly different take on the critical path. It can help you deliver your project earlier than you would otherwise, and helps prevent delays in the project finish date. The critical chain approach schedules tasks to occur as late as possible. One benefit of doing this is that you don't spend money on the project until you absolutely have to. If you have to adjust the schedule, you move tasks to occur earlier. Critical chain also focuses on resource limitations to identify the important tasks to manage. That's because resource constraints are often the toughest ones to deal with. You start by scheduling tasks with the most limited resources. So you use those people as effectively as possible. It's like filling a bucket to the brim by putting the biggest rocks in a bucket first, followed by smaller rocks and finishing up with sand. The critical chain approach uses buffers to give a project breathing room. That way, delays aren't as likely to delay the finish date. What do these buffers do? They're like adding shared time to the project. Each task doesn't get its own time buffer. Instead, sequences of tasks share a buffer. That way, only the tasks that actually need extra time use some of the buffer. You apply a couple of different types of buffers. First, you tack buffers onto the end of each sequence of tasks. Then, you add a project buffer at the end of the project to protect the overall project finish date. The critical chain approach helps deliver projects on time, or earlier than planned. 

## How to shorten a schedule

Stakeholders can be an impatient bunch. You estimate when the project will finish, and more often than not, they ask if you can deliver it sooner. Or maybe your project gets delayed and you need to make up time. Let's look at a few techniques for shortening the project schedule. 

1. Fast-tracking means you overlap tasks that normally occur one after the other. Let's say you want to finish design more quickly. You can have some folks start designing software features before the system design is complete. Fast-tracking is simple. You just add an overlap to two tasks with finish-to-start dependencies. The best tasks to fast-track are the longest tasks on the critical path. Remember, you shorten the whole project when you shorten the critical path. Fast-tracking the longest tracks on the critical path shortens the schedule and also introduces the fewest number of risks and changes. Which brings us to the disadvantage of fast-tracking. It increases risk. For example, changes to the system design could result in redoing work on the software design. 
2. The second technique, crashing, means you spend additional money to shorten the schedule, like paying for more people or expedited delivery of key materials. The critical path is still the place to look for tasks to crash. Think about it. Why would you spend money to shorten tasks if they aren't going to shorten your schedule? You also want the alternative that shortens the schedule the amount you need for the least amount of money. Start with the least expensive tasks to crash, and move onto tasks with higher price tags. Remember, you crash tasks only until you shorten the schedule by the amount you need. A crash table makes it easy to see which tasks to crash. It shows the cost to crash each critical task, and the duration you eliminate by crashing them. Choose the tasks with the lowest crash cost per week. If they have the same crash cost per week, crash the longer ones first. That way, you crash the fewest tasks. You can take crashing only so far. At some point, adding more people won't shorten the duration, because people start getting in each other's way, and on each other's nerves. New people take time to get up to speed and they can slow down existing team members who have to help them get oriented. Keep in mind, you review the critical path after every adjustment. An adjustment could change the critical path. You want to make sure the next task you work on is still on it. 
3. The third method for shortening a schedule is to cut project scope. If the tasks for deleted scope are on the critical path, cutting scope shortens the schedule. Fast-tracking, crashing, and cutting scope help shorten the project schedule. Each has its pros and cons, so choose the method that makes the most sense for the project at hand. 

## Document the baseline

Once the stakeholders have approved the project plan, it's time to define your project baseline. The baseline is a collection of the approved project documents like requirements, the schedule, budget, spread sheets, and more. Basically, everything you want to control with the change management process. With baseline documents, you can compare actual project performance to the baseline to see how the project's doing. Because the baseline is controlled by the change management process, any changes to baseline documents show up as change requests. How you define the baseline depends on what you're defining. First, save the baseline version of plan documents. If something changes, you flag those changes in a revision of the corresponding baseline document. Next, baseline the values in your project schedule. Project scheduling programs typically provide a feature for saving a baseline. The baseline includes the approved values for start and finish dates, task duration, work, costs, and so on. As you record progress or make changes to the schedule, the program will show any variances from the baseline. Once you document the project baseline, you're ready to use it to evaluate progress and project performance. 

# An Overview of Agile Project Management
 
## What is agile project management?
 
Agile projects are great when business needs change frequently, or the business wants to receive value sooner. Agile does this by frequently delivering small chunks of a product known as features. This steady release of deliverables requires a different approach to managing and implementing work. You guessed it. Agile Project Management. With Agile, work is performed in time periods called iterations or sprints. They typically last two to 12 weeks. The goals of each iteration is to deliver a complete, ready to use feature. That's how value is delivered sooner. And if business needs change, you can address that in a later iteration. Agile places more importance on people and interaction, than processes and tools. For example, people communicate when there's a reason to, not because the weekly team meeting is scheduled. To facilitate interaction, business and technical team members work in the same location or use collaboration tools to achieve the same affect. With Agile, your goal is to produce a product not reams of documentation. That doesn't mean no documentation. While the team is building features, documentation is limited to what they need to do their work. Documentation like User Guides are added to the features list. The customer is much more engaged in an Agile project. In the waterfall approach, the customer negotiates requirements upfront, and then waits for the end deliverables. In Agile, the customer collaborates throughout the development process, which means it's easier to meet their needs. Changes are not only expected, but welcomed. The Agile philosophy views change as a way to provide more value. With Agile, you can shift priorities quickly by adding new features in a later iteration. Agile performs many of the same activities in waterfall project management in a slightly different order. One way to think about it: the waterfall approach defines the scope of the project, and then estimates the time, cost, and quality to deliver that scope. Agile looks at time, cost, and quality as fixed elements. Then you figure out which features you can produce given those constraints. The Agile methodology described in this chapter starts with the Envision stage, where you define the goal and the boundaries of the project. You then repeat the Speculate, Explore, and Adapt stages to develop and test features. At the end of the project, you perform a Closing Stage once to wrap things up. Change control is done by prioritizing the backlog of features, and risk is managed with frequent releases. If you're familiar with Agile, you might think Agile means IT projects. It turns out, Agile works with non-IT projects, too, as long as deliverables can be produced and implemented in short periods and can be added to in the future. Agile works will for projects with changing business needs or a desire to produce value quickly. 

## The agile project management life cycle

The agile project management lifecycle is made up of five stages. 

1. Envision
2. Speculate
3. Explore
4. Adapt
5. Close. 

The envision stage is the start of your project. When you and the customer, well, envision what you're going to deliver. You define the project goal and objectives and figure out who you need on your team. You also set up the guidelines for how you'll run the project. Things like how the team works to solve problems, how you'll collaborate and communicate, what hours the team works. When envisioning is done you'll have a project definition that spells out the project customer, goal, objectives, scope, and stakeholders. At this point in the lifecycle an agile project diverges from the waterfall approach. For each iteration, you cycle through the speculate, explore, and adapt stages. 

The speculate stage is a about planning. Except that you plan in iteration, not the entire project. Part of the plan is identifying the business functions you'll address in the iteration and the features you'll develop to satisfy those requirements. You estimate the effort for each feature and you identifying or update risks related to those features. 

Just like that, you're at the explore stage. This is where the team builds the iteration's features. Remember how Agile uses interaction when it's needed? The business and technical folks also hold frequent peer reviews of features in development to provide feedback on what's been done. The team also tests what they're building so it's production ready at the end of the iteration. 

The adapt stage gives you a chance to act of the feedback you get and fix things that aren't quite right. Product owner or customer conducts a final review of the iteration's features. You hold a team meeting to evaluate what happened during the iteration. That way you can capture lessons learned and adjust the plans for future iterations. 

Once you complete the adapt stage, you repeat the speculate, explore, and adapt stages until all the features are complete. Or you've used up the time and budget allocated to the project. When the features are done you're ready for the close stage. It's time to make sure everything is done. You also produce a final set of lessons learned to use in future projects. The agile lifecycle steps through stages for each iteration, so you can start building features more quickly


## Plan an iteration

With an Agile project, you still plan things before you dive into the work. The Speculate stage is when the business and technical teams identify the features to include in the current iteration. A feature is a small function or deliverable that addresses a business need. In the hospital project, scheduling functions, user documentation, and training could be features. The typical naming convention is action followed by result. For example, "Find next available appointment time," or "Change procedure room." The first iteration Speculate stage takes a little longer than the following ones. That's because you have to identify and estimate all the features for the project, which is called the backlog. One way to build the backlog is to write each feature on a sticky note or index card. If your organization uses an Agile collaboration tool, you can add features there. When the feature list is complete, the business and technical teams review the features and prioritize them. The business team prioritizes from a functionality perspective. The technical team considers whether it makes more sense to build features in a specific order. Don't worry if the business folks identify new features or ask questions about existing ones. The Agile approach can take that in stride. Worst case, if the list gets really long, you can talk about features to push to a future project. The business and technical team members also work together to develop estimates for all features in the project. Once the prioritized feature list is approved by the customer or sponsor, you can create the Iteration, Milestone, and Release Plan. This plan outlines all the features, when they're scheduled in the iterations, and when they'll be implemented in the organization. And you're ready to start the Explore stage of the first iteration. Time passes and like any project, an Agile project doesn't necessarily follow your plan. Each time you start another Speculate stage, you have to reestimate, reevaluate, and reprioritize the features to decide which ones to include in the iteration, like features that weren't completed and are now back in the backlog, or new features that were identified since the original backlog was built, and the next features in the backlog that you want to include. The Speculate stage doesn't take long. If your iterations are three weeks long, you might spend a couple of days in the Speculate stage. You start each iteration by planning what will be done

## Manage the explore stage

Agile projects get you to the **explore state quickly**. This is where the team starts building and implementing features. At this stage, a lot of collaboration occurs between the business and technical folks. The bulk of the effort in the explore stage is working on the features assigned to the current iteration. The business and technical team members start each day with a short stand-up meeting, usually 15 minutes. While these meetings occur during all agile stages, they are key during the explore stage. Each person briefs the team on what they completed the day before, what they plan to finish today, and any help they need. If issues come up, you don't resolve them in the daily meeting. Instead, make a note of them, but work on resolving them after the meeting. The people who tackle the issue, report on it to the group in the next day's meeting. The only documentation that comes out of the meeting is adding issues to a register to track lessons learned. The team runs these meetings. That doesn't mean you, as the project manager, have nothing to do. Your main job is to remove obstacles so team members can focus on their work. During the meeting, you track progress on the features planned for this iteration. If features are behind schedule, you need to find out why and help get things back on track. If issues are to blame, add them and their solutions to the issue log. Speaking of issues, keep an eye on any that aren't getting resolved and offer to help. As with any meeting, if it starts to go off track, you can jump in to help refocus the team's attention. After the meetings, notify stakeholders of progress. Here's a key concept with the agile approach. You stop working on features when the iteration is scheduled to end, even if the features aren't done. For example, if your explore stage is four weeks, you stop at four weeks. If features are incomplete, you put the incomplete work into the backlog to address in a future iteration. On the other hand, if the team cranks out the iteration features faster than they estimated, you can start working on another feature from the backlog to fill in the rest of the time. The explore stage is mostly about building and implementing features, with short daily meetings to help keep things on track.

## Handle the adapt and close stages

You've made it to the last stage of an iteration, the adapt stage. It's a time for review and reflection, like spending New Year's Day evaluating your previous year, and resolving to do even better in the next. As you might expect, you and the team compare what was delivered to what was planned. You talk about what did and didn't work. The team might identify changes to make future iterations run better. Take time to review with the customer what's been done, and make sure that the features are working the way they should. At the same time, check that the features produced the business benefits they're supposed to. Don't forget to hold a lessons learned session to share feedback. That way, everyone can brainstorm ways to resolve issues and eliminate obstacles, plus, everyone can improve by taking advantage of the things that went well. Feedback can lead to changes in the next iteration's speculate stage. You might add or remove features from the backlog. The team could modify their estimates for features, or update the risk register. The customer might reprioritize the features in the backlog. Maybe the team composition needs tweaking based on people's availability, or the skills needed. Finally, you might modify processes that aren't working well. 

The end of an iteration is a great time to celebrate success. In addition, the adapt stage is a chance for team members to take a quick break, so they're fresh and enthusiastic about starting the next iteration. If there are more iterations in the project, the end of the adapt stage triggers the speculate stage for the next iteration. On the other hand, if the project schedule or budget has been used up, it's time for the close phase, and tying up loose ends. You reconcile the project financials. Have all the invoices been paid? Have you closed the project accounts so people don't charge to them by mistake? You also need to work with resource managers to reassign team members. As in any project, you communicate the project results to the customer and stakeholders. It's important to get closure at the end of each iteration, and the end of the project. That way, you confirm the success of this and future projects. For practice, build agendas for the review meetings you plan to hold during an iteration's adapt stage.

# While You Run the Project

## How to run a project

Finally, planning is done and the project is underway. Now all the work you did in planning comes into play as you push your project to completion. Executing a project starts with lining up the people and other resources you need to perform the work. 

Once team members are on board, they need to know their assignments, who they're working with, and how the project's going to go. That's what a **kickoff meeting** is for. The project-sponsoring customer can describe the mission and get everybody jazzed about the project. You can review the project plan and explain how things will work by communicating or handling change requests. You also set up a place to store project information, like your project plan, specifications, documents, and reports. It's called a project notebook, but these days, it's usually electronic. Lots of people need access to these files, so be sure to set your project notebook up where team members can get to it. After that, project execution represents performing all the work you identified in the work breakdown structure. Monitoring and controlling your project kick in at the same time as executing. Monitoring means collecting data about where the project stands. Since projects never stick to the plans you so carefully prepare, you have to figure out how to respond to the changes and surprises and problems that arise. Controlling is about correcting course to get your project back on track.  


## Techniques for communicating effectively

As a project manager, you communicate with people at all levels of the organization, from the customer and management team, to the people who do project work. Communication isn't telling someone something. It's about getting a message across so the other person understands it and does something about it. Good communication skills are key. With them, you and your team will get more and better work done. 

1. First, tell your audience why they should pay attention. 
2. Second, get to the point. That way, you'll make your point before they lose interest. 
3. Third, tailor your message to your audience. What's relevant to them? Use terms they understand. I managed a project where the client wanted a lot and they wanted it quickly. The team was overworked and exhausted. At meetings, management focused on projected sales and revenue from the project. Team members just slumped in their chairs, because all they heard was we want you to do even more work. When I talked to the team, I focused on how the project would benefit them. They perked up hearing that and we did make the project a success. 
4. Fourth, be positive and proactive. If you have to discuss a problem, describe what it is and then explain your plan to handle it. Here's some other tips for improving communication. Listening is as important as talking. When you're the audience, pay attention. No cell phones, no text messages, no emails. I know it's tough, but it's worth practicing. You'll get a lot more done. Watch for unspoken communication. Think body language and facial expressions. On the phone, listen for intonations in someone's voice. When you have difficult, delicate, or crucial messages, choose face to face meetings. That way, everyone can receive spoken and unspoken messages. Keep an open mind. Conversations or meetings are a waste if you aren't going to listen to what other people say. Paraphrase what you've heard, or ask the other person to do that. By putting information in your own words, you demonstrate that you do understand and give the other person a chance to correct you if necessary. 
5. Finally, use email effectively. Take time to write a clear message. It will actually save time by reducing back and forth email volleys. Make subject lines meaningful. If you're asking for action and have a deadline, put them right in the subject. Start with the point. Start with what you want and what's important, and then backfill with the detail. A good email is like a newspaper article. You get the key info right up front. Before you click Send, don't forget to proofread your messages. Spelling or editing mistakes and grammatical errors can radically change the meaning of a message. Confirm that your emails arrive. If you don't get a response, follow up with an email, call, or visit. Finally, be careful with humor. It doesn't work so well in email. Use it only with people you know well, and let them know you're doing it.


## Run effective meetings

Meetings chew up time and energy that people could be using to get things done. And because meetings include several people, they cost more than other work. But sometimes, they're the best way to work and communicate. If you hold a meeting, make it count. The first step to a productive meeting is to know what you're trying to accomplish. Identify the purpose of the meeting, and the results you need, like approval to move forward, or to resolve an issue. Create an agenda, that way you can make sure to cover every item. You can keep discussions on topic, and you'll have a fighting chance to finish your meeting on time. On the agenda, list the discussion topics. Don't forget to include your time estimate for each topic. The more people you have in a meeting, the harder it is to get things done. That's why the third step is to limit the attendees to the people you need to accomplish your goal. Also, give attendees a chance to prepare. Schedule the meeting when it works for them, and send the meeting invitation and materials ahead of time. For example, if you want stakeholders to approve a revised plan, send the meeting request and the plan before the meeting. Start and finish meetings on time, even if all attendees aren't there. Be strong. Don't backtrack when people show up late, that just reinforces their rude behavior. Facilitate the meeting to keep everyone focused on the meeting's objectives. As project manager, you have lots to do in a meeting. Ask someone else to be the facilitator. The facilitator kicks off the meeting with a brief introduction, the purpose of the meeting, the agenda topics, the attendees, and the ground rules for interaction. The facilitator can coax people to participate, or, if the discussion starts to go off track, they can wrangle it back on topic. For example, by adding a topic to the parking lot to be handled later. Finally, take good meeting notes. How else will you remember decisions, action items, and who's responsible for them? During the meeting, someone can document items on a white board, or a flip chart. If possible, appoint someone else to be the scribe. After the meeting, you can edit the notes to emphasize action items, who's responsible for them, and when they're due. Distribute the notes to attendees and anyone else who needs to know. If you run effective meetings, your projects will run more smoothly, and you're sure to win popularity contests at work. Give yourself a head start.

# Work with Teams


## Manage team resources

Team members are the ones who get things done and achieve the project objectives. You have to motivate the people who work on your project even though they usually report directly to someone else. Let's look at several methods to strengthen your working relationships and motivate your team members. Clearly communicate roles and responsibilities to your team members. You can build your relationships with them by helping them understand what they're responsible for and what you expect of them. Next, give people specific and achievable goals. Team members usually dig deeper to respond to challenges. Provide support and help remove obstacles that get in the way. If your people can't resolve issues or remove obstacles, help them in any way you can. That way, they can focus on the work they're supposed to do. Respect your people. They're crucial components to your project, because project work wouldn't get done without them. Being respectful helps you get the best your team has to offer. Also, provide feedback quickly. Providing positive reinforcement quickly helps build rapport, and people will do more of what works well. If someone does something wrong, jump in quickly to explain what they should be doing in a tactful way. Another important point: consistently tell the truth. By telling the truth, you earn people's trust even if they don't like what they hear. If you can, explain why you're making the requests or decisions that you do. Next, communicate regularly with team members. Status and lessons learned meetings are great ways to find out what people are doing and how well they're doing it. And of course, comparing their progress to the baseline shows whether they are as productive as you expected them to be. Finally, handle people problems quickly if they come up. The answer could be as easy as telling someone what they're supposed to do or explaining the effect they're having on others. It's important to be upfront while at the same time being tactful and respectful. If it turns out someone isn't qualified, think about how you can help them succeed or find someone else to help or take over the work. If this situation comes up, check your work packages and resource requests. Make sure you defined them clearly. If you have to replace someone, work with the functional manager to get someone else. If you can't replace the person, you might have to reset stakeholders' expectations about when work will get done, how much it'll cost, or maybe the quality. These techniques help build effective working relationships and motivate team members. To solidify your understanding, write up a sample of what you would say to your team members in your kickoff meeting.

## Understand team dynamics

People who work on your project may be working together for the first time. You know how it goes when people are getting to know one another. Work progresses in fits and starts while folks learn how to work together. To get high-quality results and productivity from your people, you have to help them develop into a team. Like people do, teams grow up through several stages before they're truly productive. One model that describes the typical phases that teams go through is called forming, storming, norming, and performing. Bruce Tuckman proposed it back in 1965. 

The first team stages is called forming because the individuals are just starting to form into a team. They aren't sure of their goal as a team, and they don't know who does what. During the forming stage, you as their leader have to clearly define the team's goals and give them direction. Like other youngsters, teams in the forming stage tend to resist and challenge authority, so be prepared to answer their questions and earn their respect. 

The second stage is called storming because things with teams usually get worse before they get better. As team members start to work out their relationships with each other, power struggles often occur. Storming teams have difficulty making decisions because of disagreements. The good news is that those disagreements lead to communication, which helps people grow as a team. When a team is storming, you have to help them stay focused on their goals and help them make decisions. After the storm has passed, you finally have a team. That's the norming stage. The team understands their common goal, and everyone pitches in to be successful. Teams don't always reach the performing stage, but if they do, (chuckles) it's pretty impressive. They know what they're supposed to accomplish, who's supposed to do what, and they often get things done very quickly, all without any help from you. You only have to step in if the team asks. Helping your people develop as a team will help your project run smoothly and deliver quality results. 


## How to manage virtual teams

Have you ever been on the phone trying to tell someone how to put something together when neither of you can see what the other person is doing? Project teams spread across the country or globe present the same kinds of issues. Distance and lack of face-to-face communication makes managing virtual teams a challenge. When people aren't in the same location, it can be hard to understand what they're saying. It's easy to misinterpret their meaning. And unless you can video conference, you don't get the benefit of reading body language or facial expressions. That's why you need really good communication to manage virtual teams. It's tough but not impossible to build long-distance relationships. The same interpersonal skills you use with office mates work with remote team members, too. Be clear in your communication, confirm understanding, be supportive and respectful, take care of your people and help them overcome obstacles, and don't forget to thank them and tell them when they do good work. Misunderstandings, communication issues, cultural differences, these all can create conflict. It's never easy to resolve conflict. It's even more challenging with virtual teams. Email is too impersonal for delicate discussions. Plus, it can lead to even more misunderstanding. Instead, use phone or video conferencing to work through conflict. Speaking of phone and video conferencing, figuring out when to hold a group session can be tough. Someone's usually up really late or much too early. Once you sort out the best time for calls, hold an initial session to talk about how communication will work on the team. Those are just a few of the challenges you might face with virtual teams and some tips on handling them. 

## How to manage technical teams

Technical folks understand technology inside and out, sometimes better than they seem to understand other people. They're also highly creative, so it's no surprise that managing them is a little different than managing other team members. First, it's important to help technical team members understand the project's business objectives. Technical people tend to focus on technology. They're good at it, and it's what they find interesting. When they understand the business side of the project, they can help identify potential problems and find solutions that no one else would think of. Technical types often strive for perfection. They also love to come up with cool solutions, which unfortunately, might not be in the project's scope. Make sure that the work that they're doing fits the project's business needs. The drive for perfection can also cause trouble during testing and fixing defects. While fixing a high-priority problem in one area, your technical team members might be tempted to fix other issues in the same area. That might be the right thing to do, but you need to make sure before they dive in. When you manage technical people, you have to juggle your approach as time goes on. Early on, you need to give them space to be creative and explore options. The rest of the time, you need to keep them on track. Make sure they complete their important, but less interesting tasks. Technology can seem cut and dry, but technical people are creative. Just like other types of creative people, technical types work better in particular environments. For example, some like to come in late and work until the wee hours of the morning. Some can't work with any kind of noise or distraction, while others need music to focus. Still, others are more productive when they can dress the way they like. Try to be flexible enough to give your technical team members the environment that awakens their muse. These are a few tips for managing your technical teams more effectively.  


# Monitor and Control Progress and Performance

## How to gather data

While work is underway, you need to track how things are going, what's been done, what's left to do, how long it took, and what it cost. Schedule and budget usually top the list, so data like actual dates, hours worked, and money spent are givens. The first thing you need to track progress is when tasks actually start. If tasks don't start when they're supposed to, a delay in the project finish could be on the horizon. While tasks are underway, get actual work hours or actual duration. Tracking hours is a little more effort, but it pays off with a more accurate picture of the progress and labor costs. You also need to know how much work or duration is still left to complete. That way you know how much longer tasks will take and whether your original estimate was on target. That's everything you need to track labor costs. Don't forget to gather other actual costs, like travel, fees, and training. If you can, set up an automated way to get the data you need through email an online form, or an enterprise reporting tool. It will free up time you can use to handle more important project management duties. Progress data tells you where your project stands and whether you need to wrangle your project back on track. 


## How to manage project change

During planning, you build a change management plan that describes how to handle project changes. Now the project is underway. It's time to put that plan to work. Let's say a change request comes in. You assign someone to evaluate the request, add it to the change request log, and submit it to the change review board, along with your recommendation to either approve or reject it. If the request is rejected, let the requester know. Make sure no one on the team starts working on it in anticipation of its approval. However, if the request is approved, you assign someone to keep an eye on it. That person updates the baseline documents to reflect the change. Maybe new tasks are needed, or the project finish date moves or cost increases. Tag baseline document edits with the change request number. That way, it's easy to see where modifications came from. The owner of the change request also adds status to the change request log, reports on progress, and shepherds the request through to completion. At that point, the owner makes one last update in the change request log to show the actual hours, cost, and project finish date, if it's changed. What if a change request has a big impact, I mean, really big? The change request board's authority does have limits. Here are a couple of exceptions to the change request approval process. Suppose a change request affects the project's business case. In that situation, the customer needs to approve the change request. If a change request affects other projects, take the request to your organization's executive board for approval. Emergencies are another special case. When a change request can't wait for the next change review board meeting, you need to act fast. Think about getting an emergency review board meeting, or you can ask the customer or sponsor to approve the request. Change is a fact of project life. By managing change requests, you can approve the ones that make sense and keep your project on track. For practice, describe what you would do for the change request depicted in the exercise files.

## Monitor and control risks

Your risk management plan includes the risks you monitor and the response you would use to handle them. Now that the project is underway, you need to keep your eyes peeled for those risks becoming reality. And if they do, you jump into action with the appropriate response. In the risk management plan, each risk you decided to monitor has an owner assigned. That person's job is to watch their assigned risk and regularly update its status. Let's take a look at what owning a risk means in detail. Before a risk occurs, you proactively implement risk responses. That means you take steps to avoid, mitigate or plan contingencies for the risk. Then you watch for signs that high-priority risks are unfolding. Also watch for the events that trigger a contingency plan. When a risk occurs, launch the response you planned. Then, track the results. Finally, regularly report risk status and update the risk log and detail sheet. Upgrade or downgrade risk probability or impact if situations have changed. Sometimes, risks go away as the project progresses and you can close them. Don't forget about lower priority risks. Changes in the project could bump up their probability or impact. As a team, check on them occasionally to see if they should be tracked. Also watch for possible new risks that have arisen due to changes to the project. Managing and controlling risks is fairly easy as long as you have a plan. The Exercise Files include a scenario in which a risk comes to fruition.  

## Understand earned value analysis

With earned value analysis, a project earns value by completing work. It helps you evaluate project schedule and cost performance. Government projects typically require it. Earned value analysis is helpful because project measures can be deceiving. Say 50% of your project's duration has passed and 50% of the budget has been spent. Sounds like things are right on track. But if only 25% of the work is complete, there's a problem. You have to finish 75% of the work with only 50% of the time and money left. I don't know about you, but I would not count on that happening. Earned value analysis uncovers problems like these because it looks at your schedule and budget in monetary terms over time. Of course, costs are already money-based, but you also measure work in terms of money by calculating how much it costs for people to perform the work. 

## Evaluate progress

When work is underway, you have to look at the schedule and cost to see where the project stands. Start with a Gantt chart that compares the current schedule to your baseline plan. The position of the taskbars in the timescale shows whether tasks are ahead of, behind or right on schedule. In this example, the current taskbars for implementation are scheduled later than the baseline so they're behind schedule. Variance values show the exact difference between your baseline and current schedule. The finished variance in this example, is negative 25 days. You would have to shorten the remaining schedule by 25 days to get it back on track. Homing in on problem tasks helps you identify what you need to change to keep your schedule on track. Scheduling programs include tools that provide early warning of delays. For example, look for incomplete tasks that are running behind schedule. Or tasks that should have started but haven't. Or tasks where work hours are behind schedule. In addition to the schedule, you will also need to compare project costs to what was budgeted. Cost variances are perfect for that. If the project is over budget, you need to find out why. Maybe scope creep is to blame. Work was done without adding budget for it. Or tasks may be taking longer than planned so labor costs are higher. Work might be going faster than planned, that incurs cost faster so it looks like it's over budget at that point. If the work required follows your estimate, the cost at the end will be back within budget. Once you see where the project stands and identify problems, you'll have the info you need to come up with solutions.  


## Learn how to solve problems

Solving problems and making decisions have a lot in common. If you think about it, most decisions you make as a project manager are about how you're going to solve or prevent a problem. Here are few tips for making problem solving easier. 

1. First, focus on what's important. Decisions and problem solving can take time and information, which are usually in short supply. So be sure to tackle the important decisions and problems first. You have to learn to ignore what isn't important, no matter how urgent it appears. When a problem arises, ask yourself, is it important? If the answer is no, ignore it. 
2. Next, make sure you understand what the problem is. Just like when you develop a problem statement for the project. 
3. Third, evaluate the impact of problems and prioritize the ones with the greatest effect on your project. With problems that have significant and long-lasting effects on your project, it's really important to pick the best solution. For issues like these, evaluate your options like you do with project strategy. On the other hand, sometimes doing nothing at all is the answer. Taking no action on purpose is an action. And it might be an acceptable alternative. 
4. Finally, ask for help if you need it. Others might identify solutions you wouldn't think of and can help you choose the best alternative. Problem solving is something you get to practice every day.  

## How to get a project back on track

If your project goes off the rails, it's time to figure out how to get it back on track. The approach you choose and how you get permission depend on several factors. If your project is behind schedule, you can look at techniques like fast tracking, crashing, and assigning overtime to pull the finish date earlier. When cost is over budget, you might consider less expensive resources or reducing other types of costs. As a last resort, you might think about reducing scope because it can shorten the schedule and reduce cost. Start with solutions you can authorize. That way you don't have to ask permission to make changes. For example, you might shuffle task assignments to complete work more quickly or at lower cost without affecting anyone outside the project. No need to ask for approval. Go ahead and reassign. If you don't have enough authority to implement a solution, ask your customer or stakeholders for approval. Lengthening the schedule, increasing the budget, and reducing scope usually require an okay from the customer, sponsor, or management team. When you ask for approval, present your recommendations along with the pros and cons of each one. And be ready to answer questions about the solution you recommend and the ones you ruled out. When your organization runs many projects, you might have to go beyond your project stakeholders. For example, if you need contingency funds or people from other projects, you'll probably have to ask the management team for those. No matter how well you manage your projects, they'll go off track from time to time. Don't panic. Evaluate your options and ask for help effectively.  

# Close a Project

## Close a project

When your project's deliverables are complete, you might think the project is finished. But you have a few activities to perform before you can call the project done. The most important part of closing is getting the customer to agree that the project completed successfully. You want to get acceptance in writing. The best way to do that is hold a meeting so the customer and stakeholders can sign the acceptance form. Then next closing step is to document lessons learned during the project. You can improve the performance of future projects by identifying what worked well, what didn't, and how things can be done better. Third, produce file documentation, and a close out report that sums up the project's performance. Finally it's time to close contracts, archive your project information, and transition your team to their next assignments. 


## How to obtain customer acceptance

The project isn't done until the customer says so, but that doesn't mean that they can keep asking you to do things. During closing, you run acceptance tests to prove that the project is complete. Back in planning, you fleshed out the project deliverables and defined clear and quantifiable success criteria. Now it's time to design acceptance tests to demonstrate whether the deliverables do what they're supposed to. You also document procedures for running those tests. Let's look at the hospital scheduling project. A first batch of acceptance tests might demonstrate that the system is running and that devices can connect to it. Once those tests are complete, you might run a second batch to see whether hospital data uploaded correctly into the new system. After that, more tests might show that the most common scheduling tasks can be completed efficiently. When the acceptance tests are done, hold a brief sign-off meeting to get the customer and other key stakeholders to sign the acceptance document. That means they agree that the deliverables are complete and successful. Congratulations. Obtaining customer acceptance is a big milestone and worthy of a celebration.


## Document lessons learned

On every project, some things go well, and other things could go better. Don't let that experience go to waste. By collecting lessons learned, you can repeat your successes and improve on your less than stellar performance. Let's look at ways to coax this info out of your team members, then you can document those lessons to benefit future projects. First, schedule time regularly for lessons learned. Don't wait until the end of the project to ask about lessons learned. By then your team members have already forgotten a lot of them. One approach is to add lessons learned as a topic in your status meetings. Keep your lessons learned sessions positive and productive. Start with what went right, ask each person for a tip or technique that help them in their work. Like, what saved you the most time recently? Or what was the gnarliest challenge you solved and how did you do it? Then move on to lessons learned from problems people faced. Remember to keep things positive. For example, what would you do differently next time? Ask people to talk about themselves that way they don't fall into blaming others. Give people the opportunity to be open and honest. Try scheduling meetings without managers to see if people will share information more freely. You might use an anonymous method for submitting lessons learned like a suggestion box. This approach can be helpful for sensitive issues, or when people are afraid to admit mistakes. Finally, document lessons learned. You can add them to your project notebook, but you can also set up a knowledge base like a web page with frequently asked questions. That way everyone in your company can learn what your team already knows. Lessons learned give your organization the opportunity to learn and grow with each new project. Assuming that the hospitals in the consortium share a collaboration system, develop a plan for sharing lessons learned among schedulers and IT teams.  

## How to prepare a close-out report

A closeout report is like a final status report for a project. It sums up the project results and helps share what you learned with others. Like a news article, start with the highlights, what the project accomplished and whether it was successful. Did the project deliver what it was supposed to? Was it on time and within budget? If not, what was the main reason why? Then you can expand on the project's schedule and cost. When did the project finish? If the project was significantly early or late, include the variances from your original dates and the reasons. That helps publicize ways to improve estimates for future projects. What was the final cost for the whole project? What did major portions of the project cost? Were there significant variances from the baseline? Depending on what the stakeholders want to know, you might include return on investment and other financial measures. Next, include information you want to share with others, like significant changes, issues, risks and lessons learned. Say the project didn't deliver the entire scope. In the closeout report, identify the scope that was and wasn't delivered. In the hospital scheduling project, an additional custom software module was added to scope to integrate the scheduling systems and the accounting system. For risks that occurred, describe how you handled them and the results. If risks occurred that weren't identified in the original risk management plan, point them out so they aren't missed in the future. Include significant lessons learned. For instance, the scheduling project showed that vendor procurement takes more time than was allocated in this project. Finally, include a summary of the effectiveness of your project management practices. Like lessons learned, you can describe what worked well and how you would plan and manage differently in the future. The project closeout report sums up what the project did and how it did it.  

## Learn how to close and transition projects

At the very end of a project, there are a few pesky but important details to tie up. First, close any contracts you signed. If your project included a contract with a customer, the signed acceptance form means the customer agrees that the contract is complete. Depending on the contract's terms, you might have other things to do, like support the project's deliverables, or perform a follow-up in a few months. Once the scheduling system is in operation, you'll run reports to see whether hospital performance objectives have been met. With vendor contracts, confirm that they did what they're supposed to do. Then, close those contracts as well. What if another group picks up where your project team leaves off? Make sure that the ongoing group is ready to take over when the project becomes part of day-to-day operations. For example, the hospital IT department takes over maintenance of the scheduling system, and the schedulers start using the system in their daily work. As project manager, help your team transition to their next assignments. The best approach is to give functional managers advanced notice that their people will be available. Then, the managers can find new assignments and plan the transitions. Archive all your project information. Archiving information electronically makes it easy to find and share. Depending on the technology that's available in your company, you might keep your project files on a network drive, in a database, or a document management system. Finally, close the accounts you used to bill the project costs. For most projects, you keep the financial books open for a few months after the project is complete. That way, you can process follow-on expenses like support. To prevent erroneous charges, close all the accounting codes, except for the ones related to the follow-up activities. With these tasks taken care of, your project is truly complete. And your assignment as project manager is over. Take a moment, or two, to celebrate what you've accomplished.  





 