---
title: "Rio 2016 Olympics [Zehui Bai]"
Author: Zehui Bai
Date: 2025.07.27
output:
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: scroll
    theme: cerulean
runtime: shiny
---

```{r setup, include=FALSE}
library(flexdashboard)
library(shiny)
library(tidyverse)
library(packcircles)
library(countrycode)
library(ggiraph)
library(plotly)
library(lubridate)
library(eeptools)
library(maps)
library(mapproj)
library(DT)

dat <- read_csv("Olympics_2016_Rio_Athletes.csv")

dat <- dat %>%
  mutate(country = case_when(
  nationality == 'IOA' ~ 'Independent Olympians',
  nationality == 'ISV' ~ 'United States Virgin Islands',
  nationality == 'LIB' ~ 'Lebanon',
  nationality == 'ROT' ~ 'Refugee Olympic Team',
  nationality == 'SIN' ~ 'Singapore',
  TRUE ~ countrycode(nationality, "ioc", "country.name")
  )) %>%
  mutate(continent = case_when(
  nationality == 'IOA' ~ 'Independent',
  nationality == 'ISV' ~ 'Americas',
  nationality == 'LIB' ~ 'Asia',
  nationality == 'ROT' ~ 'Independent',
  nationality == 'SIN' ~ 'Asia',
  nationality == 'KOS' ~ 'Europe',
  TRUE ~ countrycode(nationality, "ioc", "continent")
  )) %>%
  mutate(continent_id = case_when(
    continent == 'Americas' ~ 1,
    continent == 'Europe' ~ 2,
    continent == 'Asia' ~ 3,
    continent == 'Africa' ~ 4,
    continent == 'Oceania' ~ 5,
    continent ==  'Independent' ~ 6)) %>%
  mutate(total = gold + silver + bronze) %>%
  mutate(dob = date_of_birth)

cent <- function(x, year = 1940) {
  m <- year(x) %% 100
  y <- ifelse(m > year %% 100, 1900 + m, 2000 + m)
  update(x, year = y)
}

dob2 = as.tibble(cent(dat$dob))
dat <- dat %>%
  mutate(
    dob2 = cent(dob),
    age = floor(interval(start = dob2, end = mdy("08-05-2016")) /
                duration(n = 1, unit = "years"))
  )


```

The athletes {data-orientation=rows}
======================================

Row {data-height=500}
-----------------------------------------------------------------------

### Chart A

```{r, echo=FALSE, warning=FALSE}
library(tidyverse)
library(ggforce)       # for circle packing
library(ggiraph)       # for interactivity
library(packcircles)   # for circle layout

# Base data (assume head(dat) already loaded)
fig_dat1 <- dat %>% select(nationality, country, continent, continent_id)

# Count athletes per country
temp_dat1 <- dat %>% count(country, name = "number_of_athletes")

# Keep one row per country for join
fig_dat1 <- fig_dat1 %>% distinct(country, .keep_all = TRUE)

# Join athlete count with country-level info
fig_dat1 <- left_join(temp_dat1, fig_dat1, by = "country") %>%
  arrange(desc(number_of_athletes))

# Circle packing layout
packing <- circleProgressiveLayout(fig_dat1$number_of_athletes, sizetype = 'area')
fig_dat1 <- bind_cols(fig_dat1, packing)

# Create vertices for drawing circles
dat.gg <- circleLayoutVertices(packing, npoints = 100)

# Add back extra columns for plotting
dat.gg$continent_id <- rep(fig_dat1$continent_id, each = 101)
dat.gg$number_of_athletes <- rep(fig_dat1$number_of_athletes, each = 101)
dat.gg$country <- rep(fig_dat1$country, each = 101)
dat.gg$nationality <- rep(fig_dat1$nationality, each = 101)

# Define fill colors by continent
region_colors <- alpha(c(
  rgb(214,39,40, max = 255),     # Americas
  rgb(44,160,44, max = 255),     # Europe
  rgb(31,119,180, max = 255),    # Asia
  rgb(140,86,75, max = 255),     # Africa
  rgb(227,119,194, max = 255),   # Oceania
  rgb(255,127,14, max = 255)),   # Independent
  0.75)

# Plot
gg1 <- ggplot(data = dat.gg) +
  ggiraph::geom_polygon_interactive(
    aes(x = x, y = y, group = id,
        fill = as.factor(continent_id),
        tooltip = paste0(country, "\nAthletes: ", number_of_athletes),
        data_id = x)) +
  scale_fill_manual(
    labels = c("Americas", "Europe", "Asia", "Africa", "Oceania", "Independent"),
    values = region_colors) +
  geom_text(data = filter(fig_dat1, number_of_athletes > 1),  # adjust for small dataset
            aes(x = x, y = y, label = nationality, size = number_of_athletes),
            color = "white") +
  scale_size(range = c(5, 9)) +
  guides(size = FALSE, fill = guide_legend(title = "Region")) +
  labs(title = "Number of Athletes by Nationality & Region") +
  theme_void(base_size = 14) +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, face = "bold", size = 24)) +
  coord_equal()

# Render as interactive girafe
girafe(ggobj = gg1, width_svg = 14, height_svg = 8)

```

### Chart B

```{r, echo=FALSE, warning=FALSE}
# Load and prepare world map data
my_world_map <- map_data("world") %>%
  rename("country" = "region")  # Rename for consistency

# Add IOC (International Olympic Committee) 3-letter codes to map data
my_world_map <- my_world_map %>%
  mutate(nationality = countrycode(country, "country.name", "ioc"))

# Extract relevant country-nationality pairs from your main dataset
fig_dat2 <- dat %>% select(nationality, country) %>% distinct()

# Count the number of athletes per country
temp_dat2 <- dat %>%
  count(country, name = "Athletes")  # more concise than table + rename

# Merge athlete counts with nationality-country mapping
fig_dat2 <- inner_join(temp_dat2, fig_dat2, by = "country") %>%
  arrange(desc(Athletes))  # Sort by athlete count

# Merge with map data using IOC nationality code
my_dat2_combined <- left_join(my_world_map, fig_dat2, by = "nationality")

# Create choropleth map
gg2 <- ggplot() +
  geom_polygon(
    data = my_dat2_combined,
    mapping = aes(
      x = long, y = lat, group = group,
      text = paste(country.x),  # Tooltip text
      fill = Athletes,
      colour = "gray10"  # Static border color
    )
  ) +
  scale_fill_distiller(palette = 7, direction = 1) +
  scale_colour_manual(values = c("gray10")) +
  theme_bw() +
  guides(color = FALSE) +  # Hide color (border) legend
  coord_map(xlim = c(-180, 180)) +
  labs(
    title = "Global distribution of athletes",
    x = "", y = ""
  ) +
  theme(
    legend.position = "right",
    axis.line = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    panel.background = element_blank(),
    plot.title = element_text(size = 16, face = "bold")
  )

# Make the map interactive with tooltips
ggplotly(gg2, tooltip = c("text", "fill")) %>%
  hide_legend()

```

Row {data-height=430}
-----------------------------------------------------------------------

### Contents

#### **Page 1:** explores the range of athletes that competed in the Rio 2016 Olympic Games. Click and drag on the map to zoom on regions, double click to reset.  

<br>

#### **Page 2:** explores the medals and medalists of the Games. Configure the medals, nationality & sport input parameters on the sidebar to filter the underlying data.  

<br>

#### **Page 3:** explores the attributes of medalists. What are observed common attributes of Olympic medalists? Configure sport & sex input parameters on the sidebar to filter the underlying data.  

<br>

#### **Page 4:** shows the total medals table. Please note that this counts all individual medals awarded, whereas the official ranking tallies only 1 medal for a team placing in a sport.  
 

### Chart C

```{r, echo=FALSE}
library(dplyr)
library(plotly)

# Filter and prepare gender count data
fig_dat3 <- dat %>%
  # Uncomment and set the country you want to filter by, e.g., "USA"
  # filter(nationality == "USA") %>%
  count(sex, name = "count") %>%
  mutate(per = count / sum(count)) %>%
  arrange(count)

# Create the pie chart
plot_ly(fig_dat3,
        labels = ~sex,
        values = ~per,
        type = 'pie',
        textinfo = 'label+percent',
        insidetextfont = list(size = 20),
        hoverinfo = 'text',
        text = ~paste(sex, "athletes:", count),
        marker = list(
          colors = c("rgba(214,39,40,0.75)", "rgba(31,119,180,0.75)"),
          line = list(color = '#000000', width = 2)
        )) %>%
  layout(
    title = "<b>Percentage of Athletes by Gender</b>",
    font = list(size = 14),
    legend = list(title = list(text = '<b>Sex</b>'))
  )


```

The medalists {data-orientation=rows}
======================================

Input {.sidebar}
-----------------------------------------------------------------------

### Input parameters

```{r}
sidebar_dat1 <- dat %>%
  filter(total > 0)

selectInput(inputId = "my_medal",
            label = "Medals",
            choices = c("total", "gold", "silver", "bronze"),
            multiple = FALSE,
            selected = "total")

selectInput(inputId = "my_nationality",
            label = "Nationality",
            choices = sort(unique(sidebar_dat1$nationality)),
            multiple = TRUE,
            selected = unique(sidebar_dat1$nationality))

selectInput(inputId = "my_sport",
            label = "Sport",
            choices = sort(unique(sidebar_dat1$sport)),
            multiple = TRUE,
            selected = unique(sidebar_dat1$sport))

```


Row {data-height=500}
-----------------------------------------------------------------------

### Chart D

```{r, echo=FALSE}
renderGirafe({

  # Prepare the filtered dataset
  fig_dat4 <- dat %>%
    select(nationality, sport, country, sex, !!as.symbol(input$my_medal)) %>%
    rename("medal" = input$my_medal) %>%
    filter(nationality %in% input$my_nationality,
           sport %in% input$my_sport) %>%
    group_by(country, sex, nationality) %>%
    summarise(medal = sum(medal, na.rm = TRUE), .groups = "drop") %>%
    filter(medal > 0) %>%
    arrange(country)

  # Circle packing layout
  packing <- circleProgressiveLayout(fig_dat4$medal, sizetype = 'area')
  fig_dat4 <- bind_cols(fig_dat4, packing)

  dat.gg <- circleLayoutVertices(packing, npoints = 100)
  dat.gg$sex <- rep(fig_dat4$sex, each = 101)
  dat.gg$medal <- rep(fig_dat4$medal, each = 101)
  dat.gg$country <- rep(fig_dat4$country, each = 101)
  dat.gg$nationality <- rep(fig_dat4$nationality, each = 101)

  # Auto label filtering threshold
  max_medal <- max(dat.gg$medal, na.rm = TRUE)
  display_filter <- case_when(
    max_medal <= 5 ~ 0,
    max_medal <= 10 ~ 5,
    max_medal <= 15 ~ 10,
    TRUE ~ 15
  )

  # Render interactive plot
  if (requireNamespace("ggiraph")) {
    gg4 <- ggplot(dat.gg) +
      ggiraph::geom_polygon_interactive(
        aes(x = x,
            y = y,
            group = id,
            fill = sex,
            tooltip = paste0(country, " ", sex, " ", input$my_medal, " Medalists:\n", medal),
            data_id = x)) +
      scale_fill_manual(
        values = alpha(c(
          "female" = rgb(214, 39, 40, max = 255),
          "male"   = rgb(31, 119, 180, max = 255)), 0.75),
        labels = c("Female", "Male"),
        name = "Sex") +
      geom_text(data = filter(fig_dat4, medal > display_filter),
                aes(x = x, y = y, size = medal, label = nationality),
                color = "white") +
      scale_size(range = c(5, 9)) +
      labs(title = "Medals by Nationality and Sex") +
      guides(alpha = FALSE,
             size = FALSE,
             color = guide_legend(override.aes = list(alpha = 0.1)),
             fill = guide_legend(title = "Sex")) +
      theme_void(base_size = 14) +
      theme(
        legend.position = "right",
        legend.title = element_text(face = "bold", size = 22),
        legend.text = element_text(size = 20),
        plot.title = element_text(hjust = 0.5, size = 36, face = "bold")
      ) +
      coord_equal()

    girafe(ggobj = gg4, width_svg = 16, height_svg = 10)

  }
})

```

### Chart E

```{r, echo=FALSE}
#height vs weight, medals=size of points, facet gender?

renderPlotly({
  
gg5 <- (fig_dat5 <- dat %>% select(name,nationality,country,sport,sex,height,weight,!!as.symbol(input$my_medal)) %>%
  rename("medal"=input$my_medal) %>%
  filter(nationality %in% input$my_nationality) %>%
  filter(sport %in% input$my_sport) %>%
  filter(medal > 0) %>%
  ggplot(
    aes(x=weight,y=height,fill=sex,size=medal,text = paste("Name:", name, "\nSex:", sex, "\nCountry:", country, "\nHeight:", height, "m\nWeight:", weight, "kg\n",input$my_medal,"medals:", medal))) +
    geom_jitter() +
    labs(title="<b> Medalists by height vs. weight & sex <b>",x="Weight (kg)",y="Height (m)") +
    scale_fill_manual(values=alpha(c(rgb(214,39,40,max=255),rgb(31,119,180,max=255)),0.5),name="")) +
    scale_size_continuous(range=c(3,10),name="") +
  theme_light() +
  ylim(1.4,2.2) +
  xlim(30,160)

ggplotly(gg5,tooltip=c("text")) %>%
  layout(legend = list(title = list(text = "<b>Sex</b>")))

})

#add slider for medals/size
  
```

Row {data-height=400}
-----------------------------------------------------------------------

### Chart F

```{r, echo=FALSE}

renderPlotly({
  
 fig_dat6 <- dat %>% select(nationality,sport,gold,silver,bronze,total) %>%
    filter(nationality %in% input$my_nationality) %>%
    filter(sport %in% input$my_sport) %>%
    group_by(sport) %>% 
    mutate(count = n())

  fig_dat6 <- aggregate(cbind(gold,silver,bronze,total) ~ sport, data = fig_dat6, FUN = sum, na.rm = TRUE) %>% 
    arrange(desc(total)) %>%
    pivot_longer(cols=c(gold,silver,bronze),names_to="medal",values_to="count")

fig_dat6$medal <- factor(fig_dat6$medal, levels = c("gold", "silver", "bronze"))

 if(input$my_medal != "total"){
   fig_dat6 <- fig_dat6 %>%
     filter(medal == input$my_medal)
 } 

if(input$my_medal == "gold"){
  my_fill <- c(alpha("gold"))
} else if(input$my_medal == "silver"){
  my_fill <- c(alpha("grey"))
} else if(input$my_medal == "bronze"){
  my_fill <- c(alpha(rgb(0.804,0.498,0.196)))
} else {
  my_fill <- c(alpha("gold"),alpha("grey"),alpha(rgb(0.804,0.498,0.196)))
}

gg6 <- ggplot(fig_dat6,aes(x=reorder(sport,-count),y=count,fill= medal,text = paste(count, medal, "medalists"),color="gray50"))+geom_bar(stat="identity")+labs(x="Sport",y="Medals",title="<b> Medals awarded by sport <b>")+
  scale_fill_manual(
    values=my_fill) + 
  scale_colour_manual(values = c("gray50"),name="")+
  theme_bw()+
  theme(panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line( size=.1, color="darkgrey" ),
        axis.text.x = element_text(size=6))

ggplotly(gg6, tooltip=c("text")) %>% 
  hide_legend()

# ggplotly(gg6, width = 1750, height = 400, tooltip=c("text")) %>% 
#   hide_legend()

})

```

Attributes {data-orientation=rows}
======================================

Input {.sidebar}
-----------------------------------------------------------------------

### Input parameters

```{r}

sidebar_dat2 <- dat %>%
  filter(total > 0)

selectInput(inputId="my_sex",label="sex",choices=c("female","male"),multiple=TRUE,selected=c("female","male"))

selectInput(inputId="my_sport2",label="sport",choices=sort(unique(sidebar_dat2$sport)),multiple=TRUE,selected=unique(sidebar_dat2$sport))

```

Row {data-height=500}
-----------------------------------------------------------------------

### Chart G

```{r, echo=FALSE}

renderPlotly({
  
fig_dat7 <- dat %>% 
  select(sex, sport, weight, height, age, total) %>%
  filter(sex %in% input$my_sex) %>%
  filter(sport %in% input$my_sport2) %>%
  rename("weight_kg"="weight") %>% 
  rename("height_m"="height") %>%
  rename("age_years"="age")

fig_dat7 <- fig_dat7 %>%
  pivot_longer(cols=c(weight_kg,height_m, age_years),names_to="variable",values_to="metric")

fig_dat7 <- aggregate(total ~ metric + variable, data = fig_dat7, FUN = sum, na.rm = TRUE) 

fig_dat7$variable <- factor(fig_dat7$variable, levels = c("weight_kg", "height_m", "age_years"))

gg7 <-  ggplot(fig_dat7,
    aes(x=metric,y=total,fill=variable,color="black",group=1,text=paste0(variable, ": ",metric, "\nmedalists: ", total))) +
    geom_area() +
  geom_point(aes(x=metric, y=total)) +
    labs(title="<b> Distribution of medalists by weight, height & age <b>",x="",y="Medals") +
    facet_wrap(~variable,scales = "free") +
    scale_colour_manual(values = c("black"),name="") +
    scale_fill_manual(values = c(rgb(255,187,120,max=255),rgb(152,223,138,max=255),rgb(174,199,232,max=255)))+
  theme_light() +
  theme(plot.title = element_text(size=20,face="bold"),
        strip.text.x = element_text(size = 16, face = "bold"),
        strip.background = element_rect(color = "black", size = 1))

ggplotly(gg7, tooltip=c("text")) %>% 
  hide_legend() %>%
  layout(margin=list(t = 100))

})
#add filter by sex
#labels with unit

```

Row {data-height=500}
-----------------------------------------------------------------------

### Chart H

```{r, echo=FALSE}

renderPlotly({
  
fig_dat8 <- dat %>% 
  select(sex, sport, weight, height, age, total) %>%
  filter(sex %in% input$my_sex) %>%
  filter(sport %in% input$my_sport2) %>%
  filter(total > 0) %>%
  rename("weight_kg"="weight") %>% 
  rename("height_m"="height") %>%
  rename("age_years"="age")

fig_dat8 <- fig_dat8 %>%
  pivot_longer(cols=c(weight_kg,height_m, age_years),names_to="variable",values_to="metric")

fig_dat8$variable <- factor(fig_dat8$variable, levels = c("weight_kg", "height_m", "age_years"))

gg8 <- ggplot(fig_dat8,aes(x=as.factor(variable),y=metric,fill=variable)) +
  geom_boxplot() +
  facet_wrap(~variable,scales = "free") +
  labs(title="<b> Spread of medalists by weight, height & age <b>",x="",y="") +
  scale_fill_manual(values = c(rgb(255,187,120,max=255),rgb(152,223,138,max=255),rgb(174,199,232,max=255))) +
  theme_light() +
  theme(axis.text.x=element_blank(),
        plot.title = element_text(size=20,face="bold"),
        strip.text.x = element_text(size = 16, face = "bold"),
        strip.background = element_rect(color = "black", size = 1))

ggplotly(gg8) %>%
  hide_legend() %>%
  layout(margin=list(t = 100))

})

#tooltip custom for outliers, solution in shiny
#add filter by sex

```

Medals Table {data-orientation=rows}
======================================

Row {}
-----------------------------------------------------------------------

### Medals Table

```{r}

renderDataTable({
  
  fig_dat9 <- dat %>% 
  select(country, gold, silver, bronze, total)
  
  fig_dat9 <- aggregate(cbind(gold,silver,bronze,total) ~ country, data = fig_dat9, FUN = sum, na.rm = TRUE) %>% 
  arrange(desc(total))
  
  fig_dat9
    
})

```

